### 2007-04-16 paper/paper.tex
svn-r3: initial import
darcs-hash:20070416220638-e5a07-82bc7646dc30070f887d8a19c3911f21b0bb36ba.gz

\documentclass{acm_proc_article-sp}
\usepackage{ucs,textcomp,url}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    backref,
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\begin{document}
\title{MetaOCaml Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
\conferenceinfo{MetaOCaml Workshop}{October 2004, Vancouver}
\CopyrightYear{2004}
\maketitle
\begin{abstract}
...
\end{abstract}
\category{D.3.2}{Programming Languages}{Language
  Classifications}[Specialized application languages]
\category{H.3.5}{Information Storage and Retrieval}{Online Information
  Services}[Web-based services]
\section{Introduction}

Today, most successful sites are programs, not static
data~\cite{greenspun99panda}.

Web programming is difficult.

It is an application area that is naturally staged.  First, content is
prepared and placed on the server (publishing stage).  Next, a user's
browser requests the content (retrieval stage).  Finally, the browser
displays the content on the client (display stage).  There is an
opportunity for dynamic code to be executed at any of these stages.

Give an example: conference calendar.

\section{Performance}

\section{Related work}

\subsection{Server-side computation}
\label{sec:related-server}

\begin{itemize}
\item PHP~\cite{bakken04php} is a neat language.
\item JSP~\cite{mahmoud03jsp}
\item SMLserver~\cite{elsman02smlserver,elsman03web}
\item XCaml~\cite{baretta04xcaml}
\end{itemize}

See also the book by \citet{greenspun99panda}.

\subsection{Client-side computation}
\label{sec:related-client}
\begin{itemize}
\item JavaScript \cite{ecmascript99,flanagan01javascript}
\item MMM, with Caml applets, by \citet{rouaix96web}
\end{itemize}

\subsection{Validated HTML in functional language}
\begin{itemize}
\item \citet{elsman04typing}
\item \citet{wallace99haxml}
\item \citet{ohl04xhtml}
\item \citet{hosoya03xduce}
\end{itemize}

\section{Future directions}

\section{Conclusion}

\bibliographystyle{abbrvnat}
\bibliography{refs,../../cal}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2007-04-16 paper/paper.tex
svn-r8: 3 performance graphs working
darcs-hash:20070416221103-e5a07-6a161d3bde3f74dafb63244280b95da8420019a8.gz
svn-r9: As submitted to MetaOCaml, 7/31/2004
darcs-hash:20070416221210-e5a07-2d34a71e13844b226be37b57e0059adbe549a42f.gz

@@
-\documentclass{acm_proc_article-sp}
-\usepackage{ucs,textcomp,url}
+\documentclass[preprint]{acm_proc_article-sp}
+\usepackage{ucs,textcomp,url,graphicx,color,listings}
+\usepackage[bf]{caption2}
 \usepackage[numbers]{natbib}
@@
   \usepackage[
-    backref,
     pdfpagemode=None,
@@
 \input{fonts}
-\begin{document}
-\title{MetaOCaml Server Pages:}
+\def\MOC{MetaOCaml}
+\hyphenation{Meta-OCaml data-base Slash-dot}
+\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
+\newcommand{\myfig}[3]{%
+  \begin{figure}[tbp]%
+    #3%
+    \caption{#2}%
+    \label{fig:#1}%
+  \end{figure}}
+\newcommand{\screenshot}[2]{%
+  \myfig{#1}{#2}
+  {\centering%
+    \includegraphics[width=20pc,keepaspectratio]{#1}}}
+\newcommand{\gnuplot}[2]{%
+  \myfig{#1}{#2}{\input{#1}}}
+\newcommand{\mylst}[3]{%
+  \lstinputlisting[#1 label={lst:#2}, %
+    caption={\texttt{#2}.  #3}]{#2}}
+\newcommand{\lstfile}{\mylst{}}
+\newcommand{\lstfloat}{\mylst{float=htb,}}
+
+\title{\MOC{} Server Pages:}
 \subtitle{Web Publishing as Staged Computation}
@@
   \email{christopher.league@liu.edu}}
-\conferenceinfo{MetaOCaml Workshop}{October 2004, Vancouver}
+\conferenceinfo{\MOC{} Workshop}{October 2004, Vancouver}
 \CopyrightYear{2004}
+\toappear{Submitted to the \MOC{} Workshop (Vancouver,
+  2004).\\\textbf{Note from author:} Sorry, the exposition gets
+  really rough toward the end: too much hacking and not enough 
+  writing!  If the PC agrees that this is a worthwhile 
+  idea, I'll be happy to fix the paper \emph{ex post
+    facto\ldots}}
+
+\begin{document}
 \maketitle
 \begin{abstract}
-...
+  Modern dynamic web services are really computer programs.
+  Some parts of these programs run off-line, others run
+  server-side on each request, and still others run within
+  the browser.  In other words, web publishing is
+  \emph{staged computation,} either for better performance,
+  or because certain resources are available in one stage
+  but not another.  Unfortunately, the various web
+  programming languages target one stage only.  This is a
+  tremendous opportunity for multi-stage languages in
+  general, and for \MOC{} in particular.
+
+  We present the design of \MOC{} Server Pages.  Unlike
+  other languages in its genre, the embedded \MOC{} code
+  blocks may be annotated with staging information, so that
+  the programmer may safely and precisely control which
+  computation occurs in which stage.  A prototype web
+  server, written in OCaml, supports web sites with both
+  static and dynamic content.  We provide several sample
+  programs and demonstrate the performance gains won using
+  multi-stage programming.
 \end{abstract}
@@
   Services}[Web-based services]
-\section{Introduction}
 
-Today, most successful sites are programs, not static
-data~\cite{greenspun99panda}.
+\vfill
+\section{Motivation}
+\label{sec:intro}
 
-Web programming is difficult.
+In their infancy (ca.~1993), web sites were little more than
+clickable FTP sites with in-line images.  They enabled
+distribution of static documents, but did little to enhance
+collaboration or create a sense of community.  In contrast,
+today's users expect most web sites to offer sophisticated
+services and to permit endless personalization.  In 2004,
+even a strictly personal web site is expected to incorporate
+user-contributed content such as comments and ratings.
+\newpage
 
-It is an application area that is naturally staged.  First, content is
-prepared and placed on the server (publishing stage).  Next, a user's
-browser requests the content (retrieval stage).  Finally, the browser
-displays the content on the client (display stage).  There is an
-opportunity for dynamic code to be executed at any of these stages.
+To provide such features, web sites contain custom computer
+programs.  \citet{greenspun99panda}, in his insightful web
+publishing guide book, describes the situation like this:
 
-Give an example: conference calendar.
+\begin{quote}
+  You needn’t turn your site into a program just because the
+  body of material that you are publishing is changing.
+  Sites like Yahoo, for example, are sets of static files
+  that are periodically generated by programs grinding
+  through a dynamic database.  With this sort of
+  arrangement, the site inevitably lags behind the database,
+  but you can handle millions of hits a day without a major
+  investment in computer hardware, custom software, or
+  thought.\par If you want to make a collaborative site,
+  however, then at least some of your web pages will have to
+  be computer programs.  Pages that process user submissions
+  have to add user-supplied data to your web server’s disk.
+  Pages that display user submissions have to look through a
+  database on your server before delivering the relevant
+  contributions.~\cite[page 271]{greenspun99panda}
+\end{quote}
+
+In some notable ways, programming for the web is different
+from programming in other domains.  Interaction design is
+more of a challenge: in part because HTTP is stateless, but
+also because users want to hit the `back' button and to set
+bookmarks into the middle of
+a carefully designed sequence of dialogs.%
+\footnote{\citet{queinnec00influence} and
+  \citet{graunke01programming} examine language-based
+  approaches to solving this particular
+  problem.} %
+And unlike with shrink-wrapped applications, on the web it
+is possible (though not always desirable) to ``go live''
+immediately with a new fix or feature.
+
+Nevertheless, there is at least one dictum of program design
+that we cannot escape on the web: \emph{performance
+  matters.}  As a web publisher, visitors are your bread and
+butter.  But will your servers and scripts be ready for the
+day that your site is featured on prime time television, or
+on \texttt{slashdot.org}?  If tens of thousands of potential
+users drop by to find a sluggish (or dead) server, most of
+them will never return.
+
+This happens so often to sites featured on Slashdot---a
+``news for nerds'' discussion site---that it has come to be
+known as the \emph{Slashdot effect:} ``a site that might be
+designed to handle a few hundred hits per day can suddenly
+find itself handling that many a
+second.''~\cite{turner03slashdot}~ Although some ISPs have
+bandwidth limitations, Slashdot creator Rob Malda says most
+sites that fall suffer from poor planning and architecture:
+\begin{quote}
+  The problem is, if you have any sort of complicated code
+  on your page, that’s what kills people. I generally think
+  that sites are not dying because they fill their pipe, but
+  because they have poorly written code.
+
+  As a general rule, anybody who has a pretty good
+  understanding of web design, they’ve done a good job of
+  learning what information to cache. What information needs
+  to be pregenerated. So when you’re actually loading a
+  page, even if it’s a complicated page that looks dynamic
+  and custom, on the back end of that, what they’re really
+  doing is putting together a bunch of puzzle pieces that
+  have been pregenerated, and making the simplest, quickest
+  decisions they possibly can.~\cite{turner03slashdot}
+\end{quote}
+
+Both of the preceding quotes point (informally) to the idea
+of \emph{staging} the computation performed by a web
+service.  Indeed, web publishing is an application area that
+is naturally staged;\footnote{Nevertheless, it seems the
+  literature on meta-programming has not yet acknowledged
+  web services as a potential application area, although
+  \citet{sheard01accomp} does
+  mention \emph{mobile code.}} %
+to wit:
+\begin{enumerate}
+\item Content (text, images, programs, etc.) created
+  off-line is uploaded to the server---the \emph{publish}
+  stage;
+
+\item A user's browser requests content, which is
+  transferred from server to client---the \emph{serve}
+  stage; and finally
+
+\item The content is rendered within the user's
+  browser---the \emph{display} stage.
+\end{enumerate}
+At \emph{each} stage there is an opportunity for computation
+to take place.  Consider the example of a conference
+calendar, such as the one illustrated in
+figure~\ref{fig:confcal}.  After specifying your areas of
+interest (perhaps using the ACM classifications), the server
+delivers a table of matching conferences, with dates,
+locations, deadlines, and links to conference web sites.
+Events remain in the table until a few weeks after they
+occur, but deadlines that have passed are marked in red.
+You may click on any column header to change the sort order.
+The next time you visit, the server remembers your
+preferences.  Perhaps it even sends you email to remind you
+of upcoming submission and registration deadlines.
+
+\screenshot{confcal}{A simple conference calendar,
+  implemented as a PHP script that connects to a PostgreSQL
+  database.  Not as sophisticated as the service described
+  above!}
+
+Now, how might this conference service be staged?  Can
+anything be computed off-line (at the \emph{publish} stage)?
+Yes!  Since this page is probably part of a much larger site
+(whose structure does not change every day), the menus and
+other navigation aids can be laid out in advance.  We will
+not know which conferences to display until the user
+presents some identification (in the form of a `cookie'),
+but since the conference data changes infrequently, it may
+help to prepare the text of each row in advance.
+
+During the \emph{serve} stage, we look up the user's topic
+preferences, and ship out just the matching rows.  We delay
+sorting the table until the \emph{display} stage.  The user
+ought to be able to adjust the sort criteria without any
+further communication with the server.  What about marking
+past dates in red?  If this is also delayed until display,
+then the code could be cached client-side for long periods
+of time, yet still behave dynamically.
+
+At this point, we should emphasize the importance of
+\emph{profiling} in developing scalable web services.  This
+particular design for the conference calendar may not be
+optimal, depending on the relative speeds of the CPU,
+memory, database, network, and disk.  Rather, our aim is to
+provide a single language in which the various staging
+possibilities can be expressed naturally.
+
+This is in stark contrast to the status quo, where each
+system targets one stage only.  The Website Meta Language
+(WML) \cite{engelschall-wml} is an ``off-line HTML
+generation toolkit'' designed for the \emph{publish} stage.
+But many other programs (and countless ad-hoc scripts) spit
+out HTML pages: think of \LaTeX2HTML.  Google
+reports surprisingly many
+programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
+for creating family tree web sites from genealogy database
+files; these count as publish-stage tools.
+
+The \emph{serve} stage is well-served by the ``server page''
+languages, including JSP~\cite{sun-jsp},
+ASP~\cite{microsoft-asp},
+MSP~\cite{elsman03web}, and
+PHP~\cite{bakken04php}.  The CGI protocol~\cite{w3c-cgi}
+addresses the \emph{serve} stage, as do the embedded
+interpreters (such as \texttt{mod\_perl} and
+\texttt{mod\_python} for Apache) that exist to ameliorate some
+of the overhead of CGI.
+%%
+There is, relatively, a paucity of languages that operate
+client-side (\emph{display} stage), probably due to the
+difficulty of securing an installed base of interpreters.
+JavaScript~\cite{netscape-javascript} and Java applets are
+notable exceptions.
+
+Imagine implementing the conference calendar, as conceived
+above, using currently deployed technology: A Perl script
+outputs a PHP page which contains JavaScript!  Values are
+passed from one stage to the next as strings, and the
+programmer must manage all the quoting and persistence
+issues by hand.
+
+We present ``\MOC{} server pages,'' a new domain-specific
+language for web applications programming.  It leverages the
+staging annotations of \MOC~\cite{calcagno03meta,taha00metaml}
+to provide safe and precise control over the each stage of the
+computation.  We have implemented the system in two parts: our
+translator transforms the server page language into a \MOC{}
+module, which then can be incorporated into our multi-threaded
+HTTP 1.1 server (also written in \MOC{}).  The scalability
+gained by staging certain applications is stunning.
+
+The next section sketches the design and translation of \MOC{}
+server pages, and section~\ref{sec:eg} includes some non-trivial
+examples.  The server implementation is decribed in
+section~\ref{sec:impl}.  Performance and scalability are
+discussed in section~\ref{sec:perf}
+
+\section{Design}
+\label{sec:design}
+
+The general idea of a \emph{server page} language is that we
+  write HTML by default, and embed code %
+\lstinline{<?like this?>}.  %
+This is the standard XML syntax for \emph{processing
+instructions.}  In our case, the instructions are written in
+\MOC{}.   Here is your first \MOC{} server page:
+
+\lstinputlisting{first.meta}
+and its output:
+\begin{quote}
+  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
+\end{quote}
+If a code block raises an exception, the message is sent to the
+client's browser in bold-face, and the rest of the page is
+aborted.
+
+\def\cd#1{\lstinline{#1}}
+\def\id#1{\texttt{#1}}
+
+A very common use of code blocks is to print out (i.e., send to
+the browser) the result of evaluating some expression.  The
+syntax \cd{<?= e?>} is designated for this task; \cd{e} must
+have type \id{string}.  Alternatively, messages may be formatted
+with \id{printf} by placing the format string immediately after
+the code delimiter: \lstinputlisting{pi.meta}
+\begin{quote}
+  ........ $\pi$ ÷ 008 is 0.3927
+\end{quote}
+In a departure from the standard OCaml syntax for \id{printf}, the
+arguments are comma-delimited.
+%%
+One more kind of code delimiter is used for \emph{declarations;}
+these are evaluated during the publishing stage:
+\lstinputlisting{perm.meta}
+\begin{quote}
+Permissions on current directory are 0755
+\end{quote}
+
+Once published, this output will never change!  The \id{stat}
+call is executed only once, not on each request.
+
+To see how all this works, let's consider how a \MOC{} server
+page is translated into a proper \MOC{} program, to be executed
+at publish time.  Since the program is executed before any
+browser has requested the page, it cannot directly return or
+output HTML.  Instead, it will construct and return a \id{code}
+object which is subsequently run on each request (serve stage).
+Listing~\ref{lst:trans.meta} shows a sample \MOC{} server page,
+and listing~\ref{lst:trans.ml} contains its translation.
+
+\lstfloat{trans.meta}{This file demonstrates the various
+  kinds of code blocks.}%
+\begin{lstlisting}[language={[Objective]Caml},
+  label={lst:trans.ml}, float=htb, 
+  caption={\texttt{trans.ml}.  This is the translation of the
+    page in listing~\ref{lst:trans.meta}.  Declarations have
+    been lifted, and everything else appears within
+    \lstinline{.<brackets>.}}]
+module Trans = struct
+  let lift x = .<x>.
+  (* Declarations *)  
+  (* More declarations *) 
+  let page  = .<fun req puts -> (*@\label{line:page}@*)
+    let arg = Request.arg req in 
+    (* Code *) ;
+    puts ( (* String to be printed *) );
+    puts "Regular text.\n";
+    Printf.kprintf puts "format string" ( a )  ( b )  ( c ) ;
+    let x = (* Expression *) in (*@\label{line:in}@*)
+    (* More code *) ;
+    puts "Bye!\n";>.
+end
+\end{lstlisting}
+
+Declaration blocks have been lifted to the top; any side effects
+contained there are executed when the page is published.  The
+constructed code begins on line~\ref{line:page} of
+listing~\ref{lst:trans.ml}.  It is a function of two arguments:
+\cd{req} encapsulates the HTTP request, including the headers
+and query arguments; \cd{puts} is a function with type
+\cd{string->unit}; it transmits text across the network to the
+visitor's browser.  \cd{arg} looks up an
+argument in the query string of the request; its type is
+\cd{string->string option}.  These are all \emph{serve} (second)
+stage values.  It makes no sense to access them in the
+\emph{publish} (first) stage, and indeed the \MOC{} type system
+prevents this.
+
+A few finer points to note: the translator discards newlines that
+immediately follow code blocks (otherwise,
+listing~\ref{lst:trans.ml} would be dotted with %
+\lstinline{puts "\n"} statements).  It automatically follows
+code blocks with either a semi-colon or the \lstinline{in}
+keyword, as needed.
+
+Now that we understand how the server page is assembled into a
+staged program, the effects of adding \MOC{} staging operators
+to our pages should be predictable.  Below is a reformulation of
+the permissions example, using escape \cd{.~} and \cd{lift} to
+ensure that \id{stat} is called only once.
+\lstinputlisting{perm2.meta}
+%%
+Still, the \id{printf} conversion is performed at serve time, so
+maybe it is best to pre-generate that as well:
+\lstinputlisting{perm3.meta}
+Now, all that remains to execute at serve time is %
+\cd{puts "0755"}.  Finally it is possible to splice in
+additional code at publish time, such as in this count-down
+example:
+\lstinputlisting{count.meta}
+It removes the loop overhead by expanding to a sequence of 99
+\cd{puts} statements.
+
+These tiny examples illustrate some of the common uses of the
+escape operator, for which we developed syntactic sugar:
+
+\begin{tabular}{l@{\quad$\leadsto$\quad}l}
+\cd{<?~ a ?>} &
+\cd{<? .~( a ) ?>}
+\\
+\cd{<?~= b ?>} &
+\cd{<?= .~( b ) ?>}
+\\
+\cd{<?~let x = c ?>} &
+\cd{<?let x = .~( c ) ?>}
+\\
+\cd{<?~"fmt" d, e, f?>} &
+\cd{<?"fmt" .~(d), .~(e), .~(f)?>}
+\end{tabular}
+
+where the \cd{a} has type \cd{unit code}; \cd{b} has type
+\cd{string code}; and the types of \cd{d,e,f} match the
+format string.
+
+One last feature in the design of \MOC{} server pages is the ability
+to take publish-stage arguments.  The parameter names are specified to
+the translator in a \emph{pragma} declaration, which can appear
+anywhere in the code:
+\begin{lstlisting}[numbers=none]
+  <?pragma args x y z ?>
+\end{lstlisting}
+The translator inserts these as page parameters, like this:
+\begin{lstlisting}[numbers=none, language={[Objective]Caml}]
+  let page x y z = .< ... >.
+\end{lstlisting}
+Now the code can be instantiated with early parameters in countless
+ways.  For example, requesting the URI \id{/longcount} could print a
+count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
+are generated from the same server page.  Mapping from URIs to
+instantiated \MOC{} code is, for now, an implementation detail
+(see \S\ref{sec:impl}).
+
+\section{Examples}
+\label{sec:eg}
+
+To explore the expressiveness of this design, we now look at a
+series of web services, organized to comprise a small web
+site.\footnote{I hope to make the demo available on the web in time
+  for the workshop; this will require convincing the I.T. folks to
+  open up a non-standard port through the firewall.  For now, it may
+  be downloaded and run locally.} %
+To give the services a similar look, we developed a site-wide style
+sheet, and an OCaml function to generate a navigation bar from a
+hierarchical list of page titles and links.  The follow bindings are
+available site-wide:
+\begin{lstlisting}[numbers=none]
+  val preamble: string -> string
+  val navbar: string -> string
+  val postamble: string
+\end{lstlisting}
+The first function generates the standard XHTML header given the
+title, while the second function generates the navigation bar given
+the URI denoting the current page.  These can be called at either
+stage.  Naturally, if the title includes a dynamic argument, then
+\cd{preamble} will be delayed until the serve stage.
+
+\subsection{The ubiquitous power function}
+\label{sec:eg:power}
+
+Judging from its prevalence in the literature, we are certain that
+millions of grateful users would subscribe to an online service
+capable of computing the \emph{power} function!  The screen shot in
+figure~\ref{fig:power127} illustrates how it works.  The navigation
+bar shows the exponents for which code has been pre-generated.  It is
+not possible, with the current implementation, for the user to request
+other exponents (see \S\ref{sec:impl} for an explanation).  After
+selecting the exponent, the user types the base into the form and
+presses return.  The result is computed using OCaml's
+arbitrary-precision \id{Num} module.  The implementation of this
+service appears in listing~\ref{lst:power.meta}.
+
+\screenshot{power127}{Web browser displaying the result of
+  7\textsuperscript{127}.  The URI \texttt{/power127} maps to code
+  where the early argument (the exponent) has been set to 127, and the
+  late argument (\texttt{x}, the base) is taken from the query
+  string.}
+
+\lstfile{power.meta}{This implements the staged power function as a
+  web service.  The exponent \cd{y} is an early (publish stage)
+  argument; but the base \cd{x} is late (serve stage).}
+
+The first function wraps a string into a fixed width block.  The power
+function itself begins on line~\ref{line:power-fun}.  No output should
+
+occur until the preamble is generated on
+line~\ref{line:power-preamble}.  The user's input is translated to a
+number relatively late (line~\ref{line:power-num}); if there is a
+problem here, the navigation bar and form will have already been
+output before the exception message.
+
+In section~\ref{sec:perf}, we will measure the impact of staging on
+the scalability of this service.  To derive the un-staged version for
+comparison, we simply remove all the brackets and escapes from the
+code in listing~\ref{lst:power.meta}.
+
+\subsection{Directory browsing}
+\label{sec:eg:browse}
+
+Many web servers can be configured to permit clients to browse
+directories.  The web server generates, on the fly, an HTML page
+containing the names and attributes of (and hyperlinks to) all
+the files in the directory.  In Apache, the
+\texttt{mod\_autoindex} module provides this feature.
+
+ViewCVS is a more complex example of the same idea.  It is a
+Python program that sits on the web server and allows remote
+users to browse a CVS repository with a standard web browser.
+
+This kind of service can be fairly intensive; a single HTTP
+request is likely to generate dozens of system calls and disk
+accesses.  If the directory is viewed more often that it is
+changed, then it makes sense to cache or pre-generate the pages.
+It's simple enough to write a script to generate directory pages
+off-line, but what if we also want dynamic user-controlled
+behavior, such as sorting and filtering?
+
+Our contribution to the genre not only lists files in a given
+directory, but displays their MD5 checksums, renders their sizes
+in human-readable form (140k or 1.2M), and colorizes their names
+based on their extension or kind.  At serve time, the user may
+specify a regular expression for filtering, and one of 5
+criteria for sorting.  The screen shot in
+figure~\ref{fig:server-dir} shows the result of browsing the
+source directory of the server.
+
+\screenshot{server-dir}{Browsing the server source code directory.}
+
+This service is a bit like the conference calendar outlined in
+section~\ref{sec:intro}.  Everything that can be pre-generated
+\emph{is.}  At serve time, we just compile the regular expression,
+then filter and sort the data (unless the default order was chosen).
+The text of each row is ready and waiting.  At serve time, the files
+need not be opened or even \id{stat}ted.  The crux of the
+implementation is shown in listing~\ref{lst:dir.meta}:
+
+\mylst{lastline=39,}{dir.meta}{Implementation of directory
+  browsing service.  This is the page part; some top-level
+  declarations were omitted.  See also
+  listing~\ref{lst:dir.meta-2}.}
+
+We omitted some of the more mundane functions on this page, but
+\cd{list_files} is important, so it appears in
+listing~\ref{lst:dir.meta-2}.  In the first stage, this reads
+all the filenames in the directory and sorts them by name.  Then
+it looks up the fileinfo, computes the MD5 sum, and prepares the
+text of each row.  Finally, it generates code that matches the
+filename against a (future-stage) regular expression, and
+includes it in a list only if it matches.  In the next stage,
+this list will be constructed and sorted (if necessary).
+
+\mylst{}{dir.meta-2}{This top-level function transforms a list
+  of filenames into a code sequence that tests each name against
+  a regexp, and adds it to the list only in the case of a
+  match.}
+
+\subsection{Server introspection}
+
+Some web servers can be configured to display their status in
+response to certain URIs (such as \id{/server-status} on
+Apache).  We programmed a few status services in \MOC.  The
+screen shot in figure~\ref{fig:gc} displays garbage collection
+statistics from the OCaml \id{GC} module.
+
+\screenshot{gc}{Garbage collector statistics.}
+
+The only complicated bit in the code is to format the numbers
+using commas every three digits.  The descriptions are copied
+directly from the \id{gc.mli} interface in OCaml.  The only
+thing pre-generated here is the navigation bar (line 2).
+
+\mylst{lastline=30,}{gc.meta}{...}
+
+\section{Implementation}
+\label{sec:impl}
+
+\textit{Ugh\ldots out of time\ldots but I will explain one
+  critical detail\ldots}
+
+Currently, the \MOC{} implementation is not capable of reading
+and writing \id{code} values to a file.  This means that
+generated server pages (after the publish stage) must already be
+in the server memory when it starts.  This is an unfortunate
+limitation, as it means we must restart the entire web server to
+re-publish any particular page!
+
+A generated server page has this type:
+\begin{lstlisting}[numbers=none]
+  type code = Request.req -> (string -> unit) -> unit
+\end{lstlisting}
+As the server starts, it constructs a dictionary (\cd{code
+  StringMap.t}) mapping URIs to code.  This dictionary is
+consulted on each HTTP request.  If the URI is not found there,
+then we check the file system, in case it refers to a static
+file.  If this fails, we respond with a 404 (resource not
+found).
 
 \section{Performance}
+\label{sec:perf}
+
+One of the benefits of a \emph{compiled} implementation of staged
+programming---even byte-code compiled, like \MOC{}---is that it
+permits meaningful measurement.  Here, we describe the performance
+characteristics of the prototype, focusing in particular on the
+benefits of staging the computations performed by various web
+services.
+
+The single most important datum for evaluating web server performance
+is \textit{throughput:} the number of requests successfully answered
+per unit of time.  To establish a baseline, we first tested the
+throughput of the Caml prototype HTTP server delivering chunks of
+static data of various sizes, up to 64k bytes.
+%%
+Measurements were taken on an otherwise idle 1.8GHz Intel
+Xeon\texttrademark{} workstation\footnote{with 512kB cache, 768MB RAM,
+  and Ultra160 SCSI} running Linux 2.4.  We used \texttt{ab},
+the Apache HTTP server benchmarking tool, to issue requests
+from 8 threads simultaneously for 30 seconds.\footnote{\texttt{ab}
+  version 1.3d, available from \url{http://httpd.apache.org/}, 
+  invoked like this:
+  \texttt{ab -k -t 30 -c 8 }\textit{url}}
+
+\gnuplot{static}{Throughput for static pages.  Note the logarithmic
+  scale on the \textit x axis.}
+
+The baseline results are shown in figure~\ref{fig:static}.  The
+series marked “Caml file” was served by the \texttt{FileHandler}
+module.  For the “Caml code” series, the same files were treated
+as \MOC{} Server Pages, and thus turned into one big
+\texttt{puts} statement, executed by the \texttt{CodeHandler}
+module.  The only reason the code wins in the beginning is that
+all code pages are already loaded into the server’s memory on
+startup (to work around a limitation of the prototype---see
+sections~\ref{sec:impl} and~\ref{sec:future}), but the
+\texttt{FileHandler} must read files from disk each time.
+
+Figure~\ref{fig:static} includes comparable results for PHP, the
+popular server-side computation system.\footnote{libphp4.so
+  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
+static data files the extension \texttt{.php}, so that Apache
+would treat them as PHP code, even though they have no
+\lstinline{<?code blocks?>}.
+%%
+We omitted the results for Apache serving static files, because they
+are way off scale: Apache handled an astounding 4,438 hits per second
+for the 1k file, and 1,813 for the 64k file.  Apache is heavily
+optimized for serving static files: apart from caching, it uses the
+special \texttt{sendfile()} system call for zero-copy file transfer
+from kernel space~\cite{tranter03sendfile}.  Although our prototype is
+no match for Apache on static files, the overhead for interpreting
+code seems no worse than that of PHP.
+%%
+
+Using the same methodology, we now consider the performance of
+the staged and unstaged power functions; see
+figure~\ref{fig:power}.  The throughput for the staged pages is
+about 30\% higher in the beginning, but as the exponents
+increase (again, note the log scale in the graph) the gap
+narrows.  In this program, staging removes the loop overhead,
+but the number of multiplications performed is the same either
+way.  Eventually, the cost of those multiplications dominates
+everything else.\footnote{The final result,
+  2\textsuperscript{8191} has 2,466 digits in base 10!}
+
+\gnuplot{power}{Staged vs.\@ unstaged power function.  The
+  server page computes 2 to the given power
+  (2\textsuperscript{127}, 2\textsuperscript{255}, etc.)}
+
+Finally, we look at the performance of staged and unstaged
+directory browsing; see figure~\ref{fig:browse}.  For this task,
+we created directories containing a fixed number of files with
+random data.  The average file size was 32k.  The \emph{x}~axis
+shows the number of files in the directory.
+
+\gnuplot{browse}{Staged vs.~unstaged directory browsing.  *The
+  Apache results refer to \texttt{mod\_autoindex}; it does
+  \emph{not} compute the MD5 sum.}
+
+Directory browsing was the most realistic of the examples, and
+the benefit of staging is crystal clear.  In browsing a
+directory of 64 files, the unstaged program barely answered 18
+requests per second.  It would certainly succumb to the Slashdot
+effect!
+
+This is exactly the kind of page where the real work needs to be
+done in advance.  But that doesn't mean it needs to be a
+completely static page, either.  By carefully staging the
+computation, we gathered the file information in advance, yet
+still filtered and sorted the results on demand.  The staged
+directory browser answered more than 550 requests (for the same
+64-file directory) per second.
+
+Figure~\ref{fig:browse} also includes the performance of
+Apache's directory browser, \texttt{mod\_autoindex}.  It cannot
+keep up with the staged browser, even though it doesn't compute
+the MD5 sum or filter the files on demand.
 
 \section{Related work}
+\label{sec:related}
 
-\subsection{Server-side computation}
-\label{sec:related-server}
+We already referenced the other systems for server-side
+computation, but we would be remiss without mentioning
+XCaml~\cite{baretta04xcaml}\ldots
+
+For client-side computation, besides JavaScript and Java
+applets, there is the work by \citet{rouaix96web} on Caml
+applets.  It may be possible to update his MMM browser for use
+with \MOC{}, but we would prefer to use a plug-in approach that
+works with mainstream browsers\ldots
+
+\citet{elsman04typing}, \citet{wallace99haxml},
+\citet{hosoya03xduce}, and \citet{ohl04xhtml} validate HTML code
+in functional languages.  We could probably use OCaml's type
+system to greater advantage this way, but it may require
+validating the text parts of the server page with a more
+mainstream XML parser.
+
+\newpage
+\section{Future directions}
+\label{sec:future}
 
 \begin{itemize}
-\item PHP~\cite{bakken04php} is a neat language.
-\item JSP~\cite{mahmoud03jsp}
-\item SMLserver~\cite{elsman02smlserver,elsman03web}
-\item XCaml~\cite{baretta04xcaml}
+
+\item This really needs to be a module for a real server, such
+  as Apache or AOLserver.
+
+\item Need more serious case studies, where one logical service
+  is implemented across a series of pages.
+
+\item Need to address the major implementation shortcoming, of
+  loading all the generated pages into memory on startup.
+
+  Marshalling closures works partly in OCaml, but only if they
+  are part of the ``original program''---that's probably too
+  limiting for our case.  The abstract syntax representation of
+  code values could be dumped to a file, but the `persistent'
+  functions it refers to will need to be stored somewhere\ldots
+
 \end{itemize}
 
-See also the book by \citet{greenspun99panda}.
+\section{Conclusion}
+\label{sec:concl}
 
-\subsection{Client-side computation}
-\label{sec:related-client}
-\begin{itemize}
-\item JavaScript \cite{ecmascript99,flanagan01javascript}
-\item MMM, with Caml applets, by \citet{rouaix96web}
-\end{itemize}
+Web publishing is an important application domain that is quite
+naturally staged.  Web programmers write staged programs every
+day, but they do it the old-fashioned way: one script generates
+another as a string.  This is a tremendous opportunity for
+multi-stage languages.
 
-\subsection{Validated HTML in functional language}
-\begin{itemize}
-\item \citet{elsman04typing}
-\item \citet{wallace99haxml}
-\item \citet{ohl04xhtml}
-\item \citet{hosoya03xduce}
-\end{itemize}
+We presented the design of \MOC{} server pages, a new
+domain-specific language for web applications programming.  It
+leverages the staging annotations of \MOC{} to provide safe and
+precise control over the each stage of the computation.
 
-\section{Future directions}
-
-\section{Conclusion}
+We have shown the substantial benefits of this approach in terms
+of performance and expressiveness, although the prototype
+implementation suffers some limitations because it is unable to
+read and write generated code to a file.
 
 \bibliographystyle{abbrvnat}
+\raggedright
 \bibliography{refs,../../cal}

### 2007-04-16 paper/paper.tex
svn-r18: Changes to support re-executing stage 1 once server is live.
darcs-hash:20070416222509-e5a07-6b6a7f694ec557c1730173da3206b03b08dd89da.gz

@@
 
-\screenshot{confcal}{A simple conference calendar,
-  implemented as a PHP script that connects to a PostgreSQL
-  database.  Not as sophisticated as the service described
-  above!}
+\screenshot{confcal}{A simple conference calendar, implemented as a
+  PHP script that connects to a PostgreSQL database.  Not as
+  sophisticated as the service described in the text.}
 
@@
 \lstinline{<?like this?>}.  %
-This is the standard XML syntax for \emph{processing
-instructions.}  In our case, the instructions are written in
-\MOC{}.   Here is your first \MOC{} server page:
+PHP programmers are familiar with this syntax for embedding code, but
+in our case, the code itself is written in \MOC.  Here is your first
+\MOC{} server page:
 
@@
 sheet, and an OCaml function to generate a navigation bar from a
-hierarchical list of page titles and links.  The follow bindings are
+hierarchical list of page titles and links.  The following bindings are
 available site-wide:
@@
 function itself begins on line~\ref{line:power-fun}.  No output should
-
 occur until the preamble is generated on
@@
 \label{sec:impl}
-
-\textit{Ugh\ldots out of time\ldots but I will explain one
-  critical detail\ldots}
 
@@
 validating the text parts of the server page with a more
-mainstream XML parser.
+mainstream SGML or XML validator.
 

### 2007-04-16 paper/paper.tex
svn-r19: Switched to 'lgrind' for code listings.
All figures and listings are now generated in the paper/ sub-dir,
so there is no need to set TEXINPUTS.

Legend of browse plot is fixed.

Added <?! ?> syntax for auto lifting.  This improves a few examples,
although the tag proliferation is unfortunate.

I have a start on unstaged dir browser in PHP.

darcs-hash:20070416222546-e5a07-bfb60b1d0b4b1e906b9fac6983b3d2386205cf03.gz

@@
 \documentclass[preprint]{acm_proc_article-sp}
-\usepackage{ucs,textcomp,url,graphicx,color,listings}
+\usepackage{ucs,textcomp,url,graphicx,color}
+\usepackage[leftno,lineno5]{lgrind}
 \usepackage[bf]{caption2}
@@
   \myfig{#1}{#2}{\input{#1}}}
-\newcommand{\mylst}[3]{%
-  \lstinputlisting[#1 label={lst:#2}, %
-    caption={\texttt{#2}.  #3}]{#2}}
-\newcommand{\lstfile}{\mylst{}}
-\newcommand{\lstfloat}{\mylst{float=htb,}}
+\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}
+
+\newcounter{listing}
+\setcounter{listing}{0}
+\makeatletter
+\def\ext@listing{lol}
+\def\fnum@listing{Listing\nobreakspace\thelisting}
+\makeatother
+\newcommand{\mylst}[2]{%
+  \captionof{listing}{\textbf{#1.} #2}%
+  \label{lst:#1}%
+  \lgrindfile{#1}}
 
@@
 \toappear{Submitted to the \MOC{} Workshop (Vancouver,
-  2004).\\\textbf{Note from author:} Sorry, the exposition gets
-  really rough toward the end: too much hacking and not enough 
-  writing!  If the PC agrees that this is a worthwhile 
-  idea, I'll be happy to fix the paper \emph{ex post
-    facto\ldots}}
+  2004).}
 
@@
   write HTML by default, and embed code %
-\lstinline{<?like this?>}.  %
+\cd{\Mquote{}like this\Munquote}.
 PHP programmers are familiar with this syntax for embedding code, but
@@
 \MOC{} server page:
-
-\lstinputlisting{first.meta}
+\lgrindfile{first.meta}
 and its output:
@@
 
-\def\cd#1{\lstinline{#1}}
-\def\id#1{\texttt{#1}}
-
 A very common use of code blocks is to print out (i.e., send to
 the browser) the result of evaluating some expression.  The
-syntax \cd{<?= e?>} is designated for this task; \cd{e} must
-have type \id{string}.  Alternatively, messages may be formatted
-with \id{printf} by placing the format string immediately after
-the code delimiter: \lstinputlisting{pi.meta}
+syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
+have type \cd{string}.  Alternatively, messages may be formatted
+with \cd{printf} by placing the format string immediately after
+the code delimiter: \lgrindfile{pi.meta}
 \begin{quote}
@@
 \end{quote}
-In a departure from the standard OCaml syntax for \id{printf}, the
+In a departure from the standard OCaml syntax for \cd{printf}, the
 arguments are comma-delimited.
@@
 these are evaluated during the publishing stage:
-\lstinputlisting{perm.meta}
+\lgrindfile{perm.meta}
 \begin{quote}
@@
 
-Once published, this output will never change!  The \id{stat}
+Once published, this output will never change!  The \cd{stat}
 call is executed only once, not on each request.
@@
 browser has requested the page, it cannot directly return or
-output HTML.  Instead, it will construct and return a \id{code}
+output HTML.  Instead, it will construct and return a \cd{code}
 object which is subsequently run on each request (serve stage).
@@
 
-\lstfloat{trans.meta}{This file demonstrates the various
-  kinds of code blocks.}%
-\begin{lstlisting}[language={[Objective]Caml},
-  label={lst:trans.ml}, float=htb, 
-  caption={\texttt{trans.ml}.  This is the translation of the
-    page in listing~\ref{lst:trans.meta}.  Declarations have
+\mylst{trans.meta}{This file demonstrates the various
+  kinds of code blocks.}
+
+\mylst{trans.ml}{This is the translation of the
+    page in figure~\ref{lst:trans.meta}.  Declarations have
     been lifted, and everything else appears within
-    \lstinline{.<brackets>.}}]
-module Trans = struct
-  let lift x = .<x>.
-  (* Declarations *)  
-  (* More declarations *) 
-  let page  = .<fun req puts -> (*@\label{line:page}@*)
-    let arg = Request.arg req in 
-    (* Code *) ;
-    puts ( (* String to be printed *) );
-    puts "Regular text.\n";
-    Printf.kprintf puts "format string" ( a )  ( b )  ( c ) ;
-    let x = (* Expression *) in (*@\label{line:in}@*)
-    (* More code *) ;
-    puts "Bye!\n";>.
-end
-\end{lstlisting}
+    \cd{\Mbegincode brackets\Mendcode}}
 
@@
 and query arguments; \cd{puts} is a function with type
-\cd{string->unit}; it transmits text across the network to the
+\cd{string\to unit}; it transmits text across the network to the
 visitor's browser.  \cd{arg} looks up an
 argument in the query string of the request; its type is
-\cd{string->string option}.  These are all \emph{serve} (second)
+\cd{string\to string option}.  These are all \emph{serve} (second)
 stage values.  It makes no sense to access them in the
@@
 listing~\ref{lst:trans.ml} would be dotted with %
-\lstinline{puts "\n"} statements).  It automatically follows
-code blocks with either a semi-colon or the \lstinline{in}
+\cd{puts}~\verb+"\n"+ statements).  It automatically follows
+code blocks with either a semi-colon or the {\KWfont in}
 keyword, as needed.
@@
 to our pages should be predictable.  Below is a reformulation of
-the permissions example, using escape \cd{.~} and \cd{lift} to
-ensure that \id{stat} is called only once.
-\lstinputlisting{perm2.meta}
+the permissions example, using escape \Msplice{} and \cd{lift} to
+ensure that \cd{stat} is called only once.
+\lgrindfile{perm2.meta}
 %%
-Still, the \id{printf} conversion is performed at serve time, so
+Still, the \cd{printf} conversion is performed at serve time, so
 maybe it is best to pre-generate that as well:
-\lstinputlisting{perm3.meta}
+\lgrindfile{perm3.meta}
 Now, all that remains to execute at serve time is %
@@
 example:
-\lstinputlisting{count.meta}
+\lgrindfile{count.meta}
 It removes the loop overhead by expanding to a sequence of 99
@@
 \begin{tabular}{l@{\quad$\leadsto$\quad}l}
-\cd{<?~ a ?>} &
-\cd{<? .~( a ) ?>}
+\cd{\Mquote{\Mtilde} a \Munquote} &
+\cd{\Mquote{} \Msplice( a ) \Munquote}
 \\
-\cd{<?~= b ?>} &
-\cd{<?= .~( b ) ?>}
+\cd{\Mquote{\Mtilde=} b \Munquote} &
+\cd{\Mquote{=} \Msplice( b ) \Munquote}
 \\
-\cd{<?~let x = c ?>} &
-\cd{<?let x = .~( c ) ?>}
+\cd{\Mquote{\Mtilde}let x = c \Munquote} &
+\cd{\Mquote{} let x = \Msplice( c ) \Munquote}
 \\
-\cd{<?~"fmt" d, e, f?>} &
-\cd{<?"fmt" .~(d), .~(e), .~(f)?>}
+\cd{\Mquote{\Mtilde}\id{"fmt"} d, e, f\Munquote} &
+\cd{\Mquote{}\id{"fmt"} \Msplice(d), \Msplice(e), \Msplice(f)\Munquote}
+\\
 \end{tabular}
@@
 anywhere in the code:
-\begin{lstlisting}[numbers=none]
-  <?pragma args x y z ?>
-\end{lstlisting}
-The translator inserts these as page parameters, like this:
-\begin{lstlisting}[numbers=none, language={[Objective]Caml}]
-  let page x y z = .< ... >.
-\end{lstlisting}
+
+(EXPLAIN THIS INSTEAD USING TRANS FIGURE)
+
 Now the code can be instantiated with early parameters in countless
@@
 series of web services, organized to comprise a small web
-site.\footnote{I hope to make the demo available on the web in time
-  for the workshop; this will require convincing the I.T. folks to
-  open up a non-standard port through the firewall.  For now, it may
-  be downloaded and run locally.} %
+site.
+%\footnote{I hope to make the demo available on the web in time
+%  for the workshop; this will require convincing the I.T. folks to
+%  open up a non-standard port through the firewall.  For now, it may
+%  be downloaded and run locally.} %
 To give the services a similar look, we developed a site-wide style
@@
 available site-wide:
-\begin{lstlisting}[numbers=none]
-  val preamble: string -> string
-  val navbar: string -> string
-  val postamble: string
-\end{lstlisting}
+
+\lgrindfile{navspec-sig.ml}
+
 The first function generates the standard XHTML header given the
@@
 presses return.  The result is computed using OCaml's
-arbitrary-precision \id{Num} module.  The implementation of this
+arbitrary-precision \cd{Num} module.  The implementation of this
 service appears in listing~\ref{lst:power.meta}.
@@
 
-\lstfile{power.meta}{This implements the staged power function as a
+\mylst{power.meta}{This implements the staged power function as a
   web service.  The exponent \cd{y} is an early (publish stage)
@@
 
-\mylst{lastline=39,}{dir.meta}{Implementation of directory
-  browsing service.  This is the page part; some top-level
-  declarations were omitted.  See also
-  listing~\ref{lst:dir.meta-2}.}
+\mylst{dir.meta}{...}
 
-We omitted some of the more mundane functions on this page, but
-\cd{list_files} is important, so it appears in
-listing~\ref{lst:dir.meta-2}.  In the first stage, this reads
+In the first stage, this reads
 all the filenames in the directory and sorts them by name.  Then
@@
 this list will be constructed and sorted (if necessary).
-
-\mylst{}{dir.meta-2}{This top-level function transforms a list
-  of filenames into a code sequence that tests each name against
-  a regexp, and adds it to the list only in the case of a
-  match.}
 
@@
 
-\mylst{lastline=30,}{gc.meta}{...}
+%\lgrindfile{gc.meta}
 
@@
 Currently, the \MOC{} implementation is not capable of reading
-and writing \id{code} values to a file.  This means that
+and writing \cd{code} values to a file.  This means that
 generated server pages (after the publish stage) must already be
@@
 A generated server page has this type:
-\begin{lstlisting}[numbers=none]
-  type code = Request.req -> (string -> unit) -> unit
-\end{lstlisting}
+
+\lgrindfile{codeHandler.mli}
+
 As the server starts, it constructs a dictionary (\cd{code
@@
 would treat them as PHP code, even though they have no
-\lstinline{<?code blocks?>}.
+\cd{\Mquote{}code blocks\Munquote}.
 %%
@@
 
-\begin{itemize}
-
-\item This really needs to be a module for a real server, such
+ This really needs to be a module for a real server, such
   as Apache or AOLserver.
 
-\item Need more serious case studies, where one logical service
+ Need more serious case studies, where one logical service
   is implemented across a series of pages.
 
-\item Need to address the major implementation shortcoming, of
+ Need to address the major implementation shortcoming, of
   loading all the generated pages into memory on startup.
@@
   functions it refers to will need to be stored somewhere\ldots
-
-\end{itemize}
 

### 2007-04-16 paper/paper.tex
svn-r20: Added staged dir browser in PHP, and its performance measurements.
Improved figures and listings in paper.

darcs-hash:20070416222632-e5a07-32d60fb7a0f5085fabbba0a98521b9a1b3d1c01d.gz
svn-r21: Bug fix in translator, for when comments found in let block.  Changes to paper.
darcs-hash:20070416222703-e5a07-2caa368989cb0178a7e31f84c708dbd1d028758e.gz

@@
 \documentclass[preprint]{acm_proc_article-sp}
-\usepackage{ucs,textcomp,url,graphicx,color}
+\usepackage{ucs,textcomp,url,xspace,graphicx,color}
 \usepackage[leftno,lineno5]{lgrind}
@@
 \input{fonts}
-\def\MOC{MetaOCaml}
+\def\MOC{MetaOCaml\xspace}
 \hyphenation{Meta-OCaml data-base Slash-dot}
@@
 user-contributed content such as comments and ratings.
-\newpage
 
@@
 contained there are executed when the page is published.  The
-constructed code begins on line~\ref{line:page} of
+constructed code begins on line~{???line:page} of
 listing~\ref{lst:trans.ml}.  It is a function of two arguments:
@@
 instantiated \MOC{} code is, for now, an implementation detail
-(see \S\ref{sec:impl}).
+(see section~\ref{sec:impl}).
 
@@
 series of web services, organized to comprise a small web
-site.
-%\footnote{I hope to make the demo available on the web in time
-%  for the workshop; this will require convincing the I.T. folks to
-%  open up a non-standard port through the firewall.  For now, it may
-%  be downloaded and run locally.} %
+site.\footnote{A demo is available at
+  \url{http://comsci.liu.edu/metaserv/}}
 To give the services a similar look, we developed a site-wide style
@@
 figure~\ref{fig:power127} illustrates how it works.  The navigation
-bar shows the exponents for which code has been pre-generated.  It is
-not possible, with the current implementation, for the user to request
-other exponents (see \S\ref{sec:impl} for an explanation).  After
+bar shows the exponents for which code has been
+pre-generated.\footnote{In the current implementation, it is not
+  possible for the user to request other exponents once the server is
+  running.  See section~\ref{sec:impl} for an explanation.}  After
 selecting the exponent, the user types the base into the form and
@@
 The first function wraps a string into a fixed width block.  The power
-function itself begins on line~\ref{line:power-fun}.  No output should
+function itself begins on line~{???line:power-fun}.  No output should
 occur until the preamble is generated on
-line~\ref{line:power-preamble}.  The user's input is translated to a
-number relatively late (line~\ref{line:power-num}); if there is a
+line~{???line:power-preamble}.  The user's input is translated to a
+number relatively late (line~{???line:power-num}); if there is a
 problem here, the navigation bar and form will have already been
@@
 \texttt{mod\_autoindex} module provides this feature.
+%%
+ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
+complex example of the same idea. It is a Python program that sits on
+the web server and allows remote users to browse a CVS repository with
+a standard web browser.
 
-ViewCVS is a more complex example of the same idea.  It is a
-Python program that sits on the web server and allows remote
-users to browse a CVS repository with a standard web browser.
+This kind of service can be fairly intense; one HTTP request is likely
+to generate dozens of system calls and disk accesses.  If the
+directory is viewed more often that it is changed, then it makes sense
+to cache or pre-generate the pages.  Although it is simple enough to
+write a script to generate directory pages off-line, what if we also
+want dynamic user-controlled behavior, such as sorting and filtering?
 
-This kind of service can be fairly intensive; a single HTTP
-request is likely to generate dozens of system calls and disk
-accesses.  If the directory is viewed more often that it is
-changed, then it makes sense to cache or pre-generate the pages.
-It's simple enough to write a script to generate directory pages
-off-line, but what if we also want dynamic user-controlled
-behavior, such as sorting and filtering?
+We implemented
+contribution to the genre not only lists files in a given
+directory, but displays their MD5 checksums, renders their sizes in
+human-readable form (`1.2M' rather than `1194822'), and colorizes their
+names based on their extension or kind.  At serve time, the user may
+specify a regular expression for filtering, and one of 5 criteria for
+sorting.  The screen shot in figure~\ref{fig:server-dir} shows the
+result of browsing the source directory of the server.
 
-Our contribution to the genre not only lists files in a given
-directory, but displays their MD5 checksums, renders their sizes
-in human-readable form (140k or 1.2M), and colorizes their names
-based on their extension or kind.  At serve time, the user may
-specify a regular expression for filtering, and one of 5
-criteria for sorting.  The screen shot in
-figure~\ref{fig:server-dir} shows the result of browsing the
-source directory of the server.
-
-\screenshot{server-dir}{Browsing the server source code directory.}
+\screenshot{server-dir}{Browsing the server source code directory.
+  File names are colored according to their \emph{kind,} which is
+  computed from the extension and the file type.  See also
+  listing~\ref{lst:dir.meta}.}
 
@@
 
-\mylst{dir.meta}{...}
+\mylst{dir.meta}{This implements the directory browser shown in
+  figure~\ref{fig:server-dir}.}
 
-In the first stage, this reads
-all the filenames in the directory and sorts them by name.  Then
-it looks up the fileinfo, computes the MD5 sum, and prepares the
-text of each row.  Finally, it generates code that matches the
-filename against a (future-stage) regular expression, and
-includes it in a list only if it matches.  In the next stage,
-this list will be constructed and sorted (if necessary).
+In the first stage, this reads all the filenames in the directory and
+sorts them by name.  Then it looks up the fileinfo, computes the MD5
+sum, and prepares the text of each row.  Finally, it generates code
+that matches the filename against a (future-stage) regular expression,
+and includes it in a list only if it matches.  In the next stage, this
+list will be constructed and sorted (if necessary).
 
@@
 
-Some web servers can be configured to display their status in
-response to certain URIs (such as \id{/server-status} on
-Apache).  We programmed a few status services in \MOC.  The
-screen shot in figure~\ref{fig:gc} displays garbage collection
-statistics from the OCaml \id{GC} module.
+Some web servers can be configured to display their status in response
+to certain URIs (such as \id{/server-status} on Apache).  We
+programmed a few status services in \MOC.  The screen shot in
+figure~\ref{fig:gc} displays garbage collection statistics from the
+OCaml \id{GC} module.  The only thing pre-generated here is the
+navigation bar.
 
-\screenshot{gc}{Garbage collector statistics.}
-
-The only complicated bit in the code is to format the numbers
-using commas every three digits.  The descriptions are copied
-directly from the \id{gc.mli} interface in OCaml.  The only
-thing pre-generated here is the navigation bar (line 2).
+\screenshot{gc}{Garbage collector statistics.  The descriptions are
+  copied directly from the \id{gc.mli} interface in OCaml.}
 
@@
 
-\gnuplot{power}{Staged vs.\@ unstaged power function.  The
+\gnuplot{power}{Staged versus unstaged power function.  The
   server page computes 2 to the given power
@@
 
-\gnuplot{browse}{Staged vs.~unstaged directory browsing.  *The
-  Apache results refer to \texttt{mod\_autoindex}; it does
-  \emph{not} compute the MD5 sum.}
+\gnuplot{browse}{Staged versus unstaged directory browsing, in \MOC
+  and PHP.}
 
@@
 64-file directory) per second.
-
-Figure~\ref{fig:browse} also includes the performance of
-Apache's directory browser, \texttt{mod\_autoindex}.  It cannot
-keep up with the staged browser, even though it doesn't compute
-the MD5 sum or filter the files on demand.
 
@@
 
-\newpage
 \section{Future directions}
@@
 
+\section*{Acknowledgments}
+
+Many thanks to the anonymous reviewers of the \MOC workshop, for
+helpful recommendations on both the work and the presentation.
+
 \bibliographystyle{abbrvnat}

### 2007-04-16 paper/paper.tex
svn-r24: many changes to paper; re-did unstaged versions of power and dir.
darcs-hash:20070416222907-e5a07-ca7ce5e89e572b776759e7697c439e40ad2e6db4.gz

@@
 \usepackage{ucs,textcomp,url,xspace,graphicx,color}
-\usepackage[leftno,lineno5]{lgrind}
+\usepackage[nolineno]{lgrind}
 \usepackage[bf]{caption2}
@@
   \lgrindfile{#1}}
+\newcommand{\mylstf}[2]{%
+  \begin{figure}[tbp!]%
+    \lgrindfile{#1}%
+    \captionof{listing}{\textbf{#1.} #2}%
+    \label{lst:#1}%
+  \end{figure}}
 
-\title{\MOC{} Server Pages:}
+\title{\MOC Server Pages:}
 \subtitle{Web Publishing as Staged Computation}
@@
   \email{christopher.league@liu.edu}}
-\conferenceinfo{\MOC{} Workshop}{October 2004, Vancouver}
-\CopyrightYear{2004}
-\toappear{Submitted to the \MOC{} Workshop (Vancouver,
-  2004).}
+%\conferenceinfo{\MOC Workshop}{October 2004, Vancouver}
+%\CopyrightYear{2004}
+%\toappear{Submitted to the \MOC Workshop (Vancouver,
+%  2004).}
 
@@
   tremendous opportunity for multi-stage languages in
-  general, and for \MOC{} in particular.
+  general, and for \MOC in particular.
 
-  We present the design of \MOC{} Server Pages.  Unlike
-  other languages in its genre, the embedded \MOC{} code
+  We present the design of \MOC Server Pages.  Unlike
+  other languages in its genre, the embedded \MOC code
   blocks may be annotated with staging information, so that
@@
 
-\vfill
 \section{Motivation}
@@
 
-We present ``\MOC{} server pages,'' a new domain-specific
+We present ``\MOC server pages,'' a new domain-specific
 language for web applications programming.  It leverages the
@@
 computation.  We have implemented the system in two parts: our
-translator transforms the server page language into a \MOC{}
+translator transforms the server page language into a \MOC
 module, which then can be incorporated into our multi-threaded
-HTTP 1.1 server (also written in \MOC{}).  The scalability
+HTTP 1.1 server (also written in \MOC).  The scalability
 gained by staging certain applications is stunning.
 
-The next section sketches the design and translation of \MOC{}
+The next section sketches the design and translation of \MOC
 server pages, and section~\ref{sec:eg} includes some non-trivial
@@
 in our case, the code itself is written in \MOC.  Here is your first
-\MOC{} server page:
+\MOC server page:
 \lgrindfile{first.meta}
@@
 In a departure from the standard OCaml syntax for \cd{printf}, the
-arguments are comma-delimited.
+arguments are comma-delimited.  This way, fewer parentheses are
+required when using the escape or lift operators on the arguments.
+
+One more kind of code delimiter is used for \emph{declarations;}
+these are lifted to the top of your program, and evaluated during
+the publishing stage:
 %%
-One more kind of code delimiter is used for \emph{declarations;}
-these are evaluated during the publishing stage:
 \lgrindfile{perm.meta}
@@
 
-To see how all this works, let's consider how a \MOC{} server
-page is translated into a proper \MOC{} program, to be executed
+\subsection{Review of staging annotations}
+
+\MOC augments OCaml with just three annotations, to indicate how
+programs are to be staged.  Brackets \cd{\Mbegincode like
+  these\Mendcode} construct future-stage computation.  The code within
+is not executed \emph{now,} but just returned, as a code value that can be
+run later.
+
+Within brackets, the splice or escape operator \cd{\Msplice e} may
+appear.  It interrupts the code construction to execute the expression
+\cd{e} (in the current stage) and splice its result into the
+future-stage computation.  Thus, \cd{e} is required to return code of
+the proper type.
+
+To compute and splice in a regular (non-code) value, we define a
+function \cd{{\KWfont let} lift x = \Mbegincode x\Mendcode}; this
+takes any value and turns it into code.  We use it like this:
+\cd{\Mbegincode 2 * \Msplice(lift(3+4))\Mendcode} The addition is
+performed immediately (because it is escaped), and the result is
+spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}.
+
+Finally, there is an operator \Mrun{} to execute constructed code.
+Applying it to the example above, \cd{\Mrun{} \Mbegincode 2 * 7
+  \Mendcode} produces \cd{14}.
+
+\subsection{Translation to \MOC}
+To see how all this works, let's consider how a \MOC server
+page is translated into a proper \MOC program, to be executed
 at publish time.  Since the program is executed before any
@@
 object which is subsequently run on each request (serve stage).
-Listing~\ref{lst:trans.meta} shows a sample \MOC{} server page,
+Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
 and listing~\ref{lst:trans.ml} contains its translation.
@@
 contained there are executed when the page is published.  The
-constructed code begins on line~{???line:page} of
-listing~\ref{lst:trans.ml}.  It is a function of two arguments:
-\cd{req} encapsulates the HTTP request, including the headers
-and query arguments; \cd{puts} is a function with type
-\cd{string\to unit}; it transmits text across the network to the
-visitor's browser.  \cd{arg} looks up an
-argument in the query string of the request; its type is
-\cd{string\to string option}.  These are all \emph{serve} (second)
+constructed code begins with the \cd{{\KWfont let} page} line in
+listing~\ref{lst:trans.ml}.
+
+The \cd{a b c} represent publish-stage arguments (the names are
+specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
+(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
+details, including the headers and query arguments.  \cd{puts} is a
+function, provided by the server, to 
+transmit text across the network to the user's browser.  \cd{arg} is
+then defined to look up an argument in the query string of the
+request; its type is \cd{string\to string option}.
+
+These are all \emph{serve} (second)
 stage values.  It makes no sense to access them in the
-\emph{publish} (first) stage, and indeed the \MOC{} type system
+\emph{publish} (first) stage, and indeed the \MOC type system
 prevents this.
@@
 code blocks with either a semi-colon or the {\KWfont in}
-keyword, as needed.
+keyword, as required.
 
-Now that we understand how the server page is assembled into a
-staged program, the effects of adding \MOC{} staging operators
-to our pages should be predictable.  Below is a reformulation of
-the permissions example, using escape \Msplice{} and \cd{lift} to
-ensure that \cd{stat} is called only once.
-\lgrindfile{perm2.meta}
-%%
-Still, the \cd{printf} conversion is performed at serve time, so
-maybe it is best to pre-generate that as well:
-\lgrindfile{perm3.meta}
+\subsection{Staged code blocks}
+Now that we understand how the server page is assembled into a staged
+program, the effects of adding \MOC staging operators to our pages
+should be predictable.  Below is a reformulation of the permissions
+example, using escape \Msplice{} and \cd{lift} to ensure that
+\cd{stat} is called only once.%
+\lgrindfile{perm2.meta}%
+Still, the \cd{printf} conversion is performed at serve time, so maybe
+it is best to pre-generate that as well:%
+\lgrindfile{perm3.meta}%
 Now, all that remains to execute at serve time is %
-\cd{puts "0755"}.  Finally it is possible to splice in
-additional code at publish time, such as in this count-down
-example:
-\lgrindfile{count.meta}
+\cd{puts "0755"}.  Finally it is possible to splice in serve-stage
+parameters at publish time, such as in this count-down example:%
+\lgrindfile{count.meta}%
 It removes the loop overhead by expanding to a sequence of 99
@@
 These tiny examples illustrate some of the common uses of the
-escape operator, for which we developed syntactic sugar:
+escape operator, for which we developed syntactic sugar.  The first
+set of code blocks we define use the tilde character to indicate that
+some part of the enclosed quote is escaped:
 
@@
 
-One last feature in the design of \MOC{} server pages is the ability
+It is also common to use the escape with \cd{lift}; this means we are
+computing a value immediately and splicing it into the code.  For
+these blocks, we use the \Mbang{} character:
+
+\begin{tabular}{{l@{\quad$\leadsto$\quad}l}}
+\cd{\Mquote{\Mbang} a \Munquote} &
+\cd{\Mquote{} \Msplice(lift( a )) \Munquote}
+\\
+\cd{\Mquote{\Mbang=} b \Munquote} &
+\cd{\Mquote{=} \Msplice(lift( b )) \Munquote}
+\\
+\cd{\Mquote{\Mbang}let x = c \Munquote} &
+\cd{\Mquote{} let x = \Msplice(lift( c )) \Munquote}
+\\
+\cd{\Mquote{\Mbang}\id{"fmt"} d, e\Munquote} &
+\cd{\Mquote{}\id{"fmt"} \Msplice(lift(d)), \Msplice(lift(e))\Munquote}
+\\
+\end{tabular}
+
+We will see examples of most of these blocks in section~\ref{sec:eg}.
+
+\subsection{Publish-stage arguments}
+One last feature in the design of \MOC server pages is the ability
 to take publish-stage arguments.  The parameter names are specified to
@@
 are generated from the same server page.  Mapping from URIs to
-instantiated \MOC{} code is, for now, an implementation detail
+instantiated \MOC code is, for now, an implementation detail
 (see section~\ref{sec:impl}).
@@
 \lgrindfile{navspec-sig.ml}
-
-The first function generates the standard XHTML header given the
-title, while the second function generates the navigation bar given
-the URI denoting the current page.  These can be called at either
-stage.  Naturally, if the title includes a dynamic argument, then
-\cd{preamble} will be delayed until the serve stage.
+%%
+The functions above may be invoked at any stage.  Naturally, if the
+page title includes a dynamic argument, then \cd{preamble} will have
+to be delayed until theh serve stage.
 
@@
 \label{sec:eg:power}
+
+\screenshot{power127}{Web browser displaying the result of
+  7\textsuperscript{127}.  The URI \texttt{/power127} maps to code
+  where the early argument (the exponent) has been set to 127, and the
+  late argument (\texttt{x}, the base) is taken from the query
+  string.}
+
+\mylst{power.meta}{This implements the staged power function as a
+  web service.  The exponent \cd{y} is an early (publish stage)
+  argument; but the base \cd{x} is late (serve stage).}
 
@@
 presses return.  The result is computed using OCaml's
-arbitrary-precision \cd{Num} module.  The implementation of this
-service appears in listing~\ref{lst:power.meta}.
+arbitrary-precision \cd{Num} module.  The complete, documented script
+appears in listing~\ref{lst:power.meta}.
 
-\screenshot{power127}{Web browser displaying the result of
-  7\textsuperscript{127}.  The URI \texttt{/power127} maps to code
-  where the early argument (the exponent) has been set to 127, and the
-  late argument (\texttt{x}, the base) is taken from the query
-  string.}
-
-\mylst{power.meta}{This implements the staged power function as a
-  web service.  The exponent \cd{y} is an early (publish stage)
-  argument; but the base \cd{x} is late (serve stage).}
-
-The first function wraps a string into a fixed width block.  The power
-function itself begins on line~{???line:power-fun}.  No output should
-occur until the preamble is generated on
-line~{???line:power-preamble}.  The user's input is translated to a
-number relatively late (line~{???line:power-num}); if there is a
-problem here, the navigation bar and form will have already been
-output before the exception message.
+The user's input is converted to a number relatively late in the
+script.  If \cd{num\_of\_string} generates an exception (perhaps
+because the user typed non-numeric text into the box), the navigation
+bar and form will have already been output before the error message
+appears.
 
 In section~\ref{sec:perf}, we will measure the impact of staging on
-the scalability of this service.  To derive the un-staged version for
-comparison, we simply remove all the brackets and escapes from the
-code in listing~\ref{lst:power.meta}.
+the scalability of this application.  To derive the un-staged version
+for comparison, we simply convert \cd{\Mquote{\Mtilde}} and
+\cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
+brackets and escapes from the code in listing~\ref{lst:power.meta}.
 
@@
 
+Now we consider a more substantial example.
 Many web servers can be configured to permit clients to browse
@@
 ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
-complex example of the same idea. It is a Python program that sits on
-the web server and allows remote users to browse a CVS repository with
-a standard web browser.
+complex example of the same idea; it allows remote users to browse a
+CVS repository with a standard web browser.
 
@@
 write a script to generate directory pages off-line, what if we also
-want dynamic user-controlled behavior, such as sorting and filtering?
+want dynamic behavior, such as user-controlled sorting and filtering?
 
-We implemented
-contribution to the genre not only lists files in a given
-directory, but displays their MD5 checksums, renders their sizes in
-human-readable form (`1.2M' rather than `1194822'), and colorizes their
-names based on their extension or kind.  At serve time, the user may
-specify a regular expression for filtering, and one of 5 criteria for
-sorting.  The screen shot in figure~\ref{fig:server-dir} shows the
-result of browsing the source directory of the server.
+Our implementation not only lists files in a given directory, but
+displays their MD5 checksums, renders their sizes in human-readable
+form (`1.2M' rather than `1194822'), and colors their names based on
+their extension or file type.  We also support dynamic (serve time)
+customization: the user may specify a regular expression for
+filtering, and select one of 5 criteria for sorting.  The screen shot
+in figure~\ref{fig:server-dir} shows the result of browsing the source
+directory of the server itself.
 
@@
 
-This service is a bit like the conference calendar outlined in
-section~\ref{sec:intro}.  Everything that can be pre-generated
-\emph{is.}  At serve time, we just compile the regular expression,
-then filter and sort the data (unless the default order was chosen).
-The text of each row is ready and waiting.  At serve time, the files
-need not be opened or even \id{stat}ted.  The crux of the
-implementation is shown in listing~\ref{lst:dir.meta}:
+This service is a bit like the conference calendar proposed in
+section~\ref{sec:intro}.  We pre-generate \emph{everything} that does
+not rely on serve-time parameters.  The stats and MD5 sums of the
+files are collected \emph{once,} in the publish stage.  The text of
+each possible row is prepared in advance.  Once the user's request is
+made, we filter and sort the list, then output the pre-generated text
+of each remaining row.  The files need not be opened or even
+\id{stat(2)}ed while the user is waiting.  Listing~\ref{lst:dir.meta}
+contains the complete script, with documentation.
 
@@
 
-In the first stage, this reads all the filenames in the directory and
-sorts them by name.  Then it looks up the fileinfo, computes the MD5
-sum, and prepares the text of each row.  Finally, it generates code
-that matches the filename against a (future-stage) regular expression,
-and includes it in a list only if it matches.  In the next stage, this
-list will be constructed and sorted (if necessary).
+Much of the above code is unaffected by staging; functions like
+\cd{sort\_by}, \cd{human\_size}, and even \cd{fileinfo} are oblivious
+to the stage in which they are run.  Therefore, the staged code comprises a
+relatively small portion of the entire program: mainly the function
+\cd{list\_files} and its call site.
+
+To understand the staging technique used here, it may help to examine
+the output of just the first stage of computation.  The following
+listing is cleaned up from the \MOC pretty-printer, with references to
+compiled values resolved in-place.
+
+\lgrindfile{dir-out.ml}
+
+The \cd{list} is formed by testing filenames (in reverse alphabetical
+order) against the compiled regular expression \cd{rc}, including only
+those that match.  The file info record for each file has already
+been prepared and appears in the code as a value.
+
+Because the file information is compiled into the code, any changes to
+the file system following the first stage of execution will \emph{not}
+automatically appear on the web interface.  To see the latest files,
+we need to re-run stage one.  This cannot be done from within the \MOC
+server page itself, but we have programmed the server to regenerate
+pages whenever the `\cd{!}' character is appended to the URI.  The
+directory browser pages feature a `Regenerate' button at the bottom
+which will bring them up to date by running the publish stage code
+again, and caching the result for subsequent requests.
 
@@
 
-Currently, the \MOC{} implementation is not capable of reading
+Currently, the \MOC implementation is not capable of reading
 and writing \cd{code} values to a file.  This means that
@@
 \label{sec:perf}
+One of the benefits of a \emph{compiled} implementation of staged
+programming---even byte-code compiled, like \MOC---is that it permits
+meaningful measurement.  Here, we describe the performance
+characteristics of the prototype, focusing in particular on the
+benefits of staging various web services.
 
-One of the benefits of a \emph{compiled} implementation of staged
-programming---even byte-code compiled, like \MOC{}---is that it
-permits meaningful measurement.  Here, we describe the performance
-characteristics of the prototype, focusing in particular on the
-benefits of staging the computations performed by various web
-services.
-
-The single most important datum for evaluating web server performance
+The single most important metric for evaluating web server performance
 is \textit{throughput:} the number of requests successfully answered
@@
 
-The baseline results are shown in figure~\ref{fig:static}.  The
-series marked “Caml file” was served by the \texttt{FileHandler}
-module.  For the “Caml code” series, the same files were treated
-as \MOC{} Server Pages, and thus turned into one big
-\texttt{puts} statement, executed by the \texttt{CodeHandler}
-module.  The only reason the code wins in the beginning is that
-all code pages are already loaded into the server’s memory on
-startup (to work around a limitation of the prototype---see
-sections~\ref{sec:impl} and~\ref{sec:future}), but the
-\texttt{FileHandler} must read files from disk each time.
+The baseline results are shown in figure~\ref{fig:static}.  The series
+marked “Caml file” was served by the \texttt{FileHandler} module.  For
+the “Caml code” series, the same files were treated as \MOC Server
+Pages, and thus turned into one big \texttt{puts} statement, to be
+executed by the \texttt{CodeHandler}.  The only reason the code
+wins in the beginning is that all code pages are already loaded into
+the server’s memory on startup (to work around a limitation of the
+prototype---see section~\ref{sec:impl}), but the \texttt{FileHandler}
+must read files from disk each time.
 
@@
 optimized for serving static files: apart from caching, it uses the
-special \texttt{sendfile()} system call for zero-copy file transfer
+special \texttt{sendfile(2)} system call for zero-copy file transfer
 from kernel space~\cite{tranter03sendfile}.  Although our prototype is
@@
 
-Finally, we look at the performance of staged and unstaged
-directory browsing; see figure~\ref{fig:browse}.  For this task,
-we created directories containing a fixed number of files with
-random data.  The average file size was 32k.  The \emph{x}~axis
-shows the number of files in the directory.
+Finally, we look at the performance of staged and unstaged directory
+browsing; see figure~\ref{fig:browse}.  For this task, we created
+directories containing a fixed number of files with random data.  The
+average file size was 32k.  The \emph{x}~axis shows the number of
+files in the directory.  Besides the staging factor, two
+implementation languages are compared: we coded the same functionality
+in PHP.  The `staged PHP' version must be run first from the command
+line; this outputs another PHP script which is then run by the server.
 
@@
 
-Directory browsing was the most realistic of the examples, and
-the benefit of staging is crystal clear.  In browsing a
-directory of 64 files, the unstaged program barely answered 18
-requests per second.  It would certainly succumb to the Slashdot
-effect!
+Directory browsing was the most realistic of the examples, and the
+benefit of staging is crystal clear.  In browsing a directory of 64
+files, the unstaged programs barely answered 18 requests per second.
+They would certainly succumb to the Slashdot effect!
 
-This is exactly the kind of page where the real work needs to be
-done in advance.  But that doesn't mean it needs to be a
-completely static page, either.  By carefully staging the
-computation, we gathered the file information in advance, yet
-still filtered and sorted the results on demand.  The staged
-directory browser answered more than 550 requests (for the same
-64-file directory) per second.
+This is exactly the kind of page where the real work needs to be done
+in advance.  But that doesn't mean it needs to be a completely static
+page, either.  By carefully staging the computation, we gathered the
+file information in advance, yet still filtered and sorted the results
+on demand.  The staged \MOC directory browser answered more than 550
+requests (for the same 64-file directory) per second.
 
-\section{Related work}
+The degradation in performance of the staged PHP version is most
+likely due to the fact that the stage two script must be \emph{parsed}
+on each request; the size of the script is proportional to the number
+of files in the directory.  With \MOC, the code size is also
+proportional to the number of files, but the values are already in
+memory.
+
+\section{Related and future work}
 \label{sec:related}
 
-We already referenced the other systems for server-side
-computation, but we would be remiss without mentioning
-XCaml~\cite{baretta04xcaml}\ldots
+Most of the \emph{server page} systems embed programs within web pages
+using similar techniques; exemplars include JSP~\cite{sun-jsp},
+ASP~\cite{microsoft-asp}, PHP~\cite{bakken04php},
+MSP~\cite{elsman03web} (based on SML), and XCaml~\cite{baretta04xcaml}
+(based on OCaml).  Ours appears to be the first such language with
+explicit support for \emph{staging} the computation.  We inherited the
+staging annotations of \MOC~\cite{calcagno03meta} and designed several
+new kinds of code blocks based on them.
 
-For client-side computation, besides JavaScript and Java
-applets, there is the work by \citet{rouaix96web} on Caml
-applets.  It may be possible to update his MMM browser for use
-with \MOC{}, but we would prefer to use a plug-in approach that
-works with mainstream browsers\ldots
+Unlike more ad-hoc staging methods, \MOC server pages guarantee the
+type safety for generated code up front.  We do not, however, make any
+guarantees about the validity of the generated HTML.
+\citet{elsman04typing}, \citet{wallace99haxml}, \citet{hosoya03xduce},
+and \citet{ohl04xhtml} leverage ML-like type systems to validate
+(X)HTML generators.  Integrating their ideas into \MOC server pages
+may permit validation of the generated document as well.
 
-\citet{elsman04typing}, \citet{wallace99haxml},
-\citet{hosoya03xduce}, and \citet{ohl04xhtml} validate HTML code
-in functional languages.  We could probably use OCaml's type
-system to greater advantage this way, but it may require
-validating the text parts of the server page with a more
-mainstream SGML or XML validator.
+Our server page language is defined by translation into \MOC.
+Unfortunately, this means that error messages refer to the translated
+code, not the original, embedded code.  Camlp4 is a flexible
+pre-processor capable of modifying the concrete syntax of
+OCaml~\cite{rauglaudre03camlp4} while maintaining usable error
+messages.  Formulating the server page syntax with Camlp4 would likely
+be an improvement over the current, ad-hoc translator.
 
-\section{Future directions}
-\label{sec:future}
+In motivating \MOC server pages, we observed that the computation
+associated with a web page naturally decomposed into three stages:
+publish, serve, and display.  Our language, however, was designed to
+support just the first two stages.  It is straightforward to imagine
+an extension to the third stage (since \MOC itself has no constraints
+on the number of stages) but the implementation may be tricky.  First,
+we must overcome the process boundary between server and client.  We
+currently have no way to export \MOC code blocks from the program that
+created them into another program.  (On the server, we circumvented
+this limitation by running the publish and serve stages within one
+process.)  Next, we need a browser capable of loading and executing
+\MOC code.  \citet{rouaix96web} demonstrated a browser called MMM,
+written in Caml, that could run applets loaded as Caml byte-code.  It
+may be possible to update his browser for use with \MOC, but
+developing a plug-in that works with conventional browsers would be
+preferable.
 
- This really needs to be a module for a real server, such
-  as Apache or AOLserver.
+One of the major shortcomings of our implementation is that all the
+pages must be loaded into memory when the server starts.  Ideally, a
+separate tool would translate the server page source directly to
+byte-code, stored in an image on disk.  Then the server would map URIs
+to these byte-code files, loading and executing them on demand.
+Whenever we want to re-run the publish stage, we could do so
+independently of the server.  Two limitations of the \MOC
+implementation itself appear to stand in the way of this ideal.
+First, code values cannot at present be written to or read from files.
+Second, code values remain polymorphic until they are \emph{run} with
+the \Mrun{} operator.  Therefore, they cannot be passed to functions
+or stored in data structures (at least until OCaml supports rank-2
+polymorphism).
 
- Need more serious case studies, where one logical service
-  is implemented across a series of pages.
+More substantial applications are needed to demonstrate both the
+performance gains and the expressiveness of this approach.  A web
+service that uses a database and is spread over several pages would be
+more realistic.  \citet{queinnec00influence} and
+\citet{graunke01programming} use features of functional languages to
+express user interactions more naturally, given the stateless nature
+of HTTP.  We suspect that these techniques are orthogonal to the
+staging features provided by \MOC.
 
- Need to address the major implementation shortcoming, of
-  loading all the generated pages into memory on startup.
-
-  Marshalling closures works partly in OCaml, but only if they
-  are part of the ``original program''---that's probably too
-  limiting for our case.  The abstract syntax representation of
-  code values could be dumped to a file, but the `persistent'
-  functions it refers to will need to be stored somewhere\ldots
+Finally, although our HTTP implementation stands up to the Apache
+benchmarking tool, it is not likely to win any awards for reliability
+or flexibility.  Our ideas would likely have greater impact if they
+could be implemented as a module for a real server such as Apache or
+AOLserver.
 
@@
 
-Web publishing is an important application domain that is quite
+Web publishing is an important application domain that is
 naturally staged.  Web programmers write staged programs every
-day, but they do it the old-fashioned way: one script generates
+day, but they do it the old-fashioned way: one script outputs
 another as a string.  This is a tremendous opportunity for
@@
 
-We presented the design of \MOC{} server pages, a new
+We presented the design of \MOC server pages, a new
 domain-specific language for web applications programming.  It
-leverages the staging annotations of \MOC{} to provide safe and
+leverages the staging annotations of \MOC to provide safe and
 precise control over the each stage of the computation.
-
+%%
 We have shown the substantial benefits of this approach in terms

### 2007-04-16 paper/paper.tex
svn-r25: paper updated -- as submitted to MetaOCaml workshop
darcs-hash:20070416223017-e5a07-ecd5a932f5585790b224cf5f4e386aa60cf4cb5f.gz

@@
 
-In their infancy (ca.~1993), web sites were little more than
-clickable FTP sites with in-line images.  They enabled
-distribution of static documents, but did little to enhance
-collaboration or create a sense of community.  In contrast,
-today's users expect most web sites to offer sophisticated
-services and to permit endless personalization.  In 2004,
-even a strictly personal web site is expected to incorporate
-user-contributed content such as comments and ratings.
+Modern dynamic web sites support many features for user collaboration
+and personalization.  To provide such services, web sites contain
+custom computer programs, often written in one of a family of
+programming languages that have grown up around (or been adapted for)
+the web.
 
-To provide such features, web sites contain custom computer
-programs.  \citet{greenspun99panda}, in his insightful web
-publishing guide book, describes the situation like this:
-
-\begin{quote}
-  You needn’t turn your site into a program just because the
-  body of material that you are publishing is changing.
-  Sites like Yahoo, for example, are sets of static files
-  that are periodically generated by programs grinding
-  through a dynamic database.  With this sort of
-  arrangement, the site inevitably lags behind the database,
-  but you can handle millions of hits a day without a major
-  investment in computer hardware, custom software, or
-  thought.\par If you want to make a collaborative site,
-  however, then at least some of your web pages will have to
-  be computer programs.  Pages that process user submissions
-  have to add user-supplied data to your web server’s disk.
-  Pages that display user submissions have to look through a
-  database on your server before delivering the relevant
-  contributions.~\cite[page 271]{greenspun99panda}
-\end{quote}
-
-In some notable ways, programming for the web is different
-from programming in other domains.  Interaction design is
-more of a challenge: in part because HTTP is stateless, but
-also because users want to hit the `back' button and to set
-bookmarks into the middle of
-a carefully designed sequence of dialogs.%
-\footnote{\citet{queinnec00influence} and
-  \citet{graunke01programming} examine language-based
-  approaches to solving this particular
-  problem.} %
-And unlike with shrink-wrapped applications, on the web it
-is possible (though not always desirable) to ``go live''
-immediately with a new fix or feature.
-
-Nevertheless, there is at least one dictum of program design
-that we cannot escape on the web: \emph{performance
-  matters.}  As a web publisher, visitors are your bread and
-butter.  But will your servers and scripts be ready for the
-day that your site is featured on prime time television, or
-on \texttt{slashdot.org}?  If tens of thousands of potential
-users drop by to find a sluggish (or dead) server, most of
-them will never return.
+There is at least one dictum of program design that we cannot escape
+on the web: \emph{performance matters.}  As a web publisher, visitors
+are your bread and butter.  But will your servers and scripts be ready
+for the day that your site is featured on prime time television, or on
+\texttt{slashdot.org}?  If tens of thousands of potential users drop
+by to find a sluggish (or dead) server, most of them will never
+return~\cite{greenspun99panda}.
 
@@
 
-Both of the preceding quotes point (informally) to the idea
-of \emph{staging} the computation performed by a web
-service.  Indeed, web publishing is an application area that
-is naturally staged;\footnote{Nevertheless, it seems the
-  literature on meta-programming has not yet acknowledged
+This quote points (informally) to the idea of \emph{staging} the
+computation performed by a web service.  Indeed, web publishing is an
+application area that is naturally staged:\footnote{Nevertheless, it
+  seems the literature on meta-programming has not yet acknowledged
   web services as a potential application area, although
@@
   mention \emph{mobile code.}} %
-to wit:
 \begin{enumerate}
@@
 During the \emph{serve} stage, we look up the user's topic
-preferences, and ship out just the matching rows.  We delay
-sorting the table until the \emph{display} stage.  The user
+preferences, and ship out just the matching rows.  If we delay
+sorting the table until the \emph{display} stage, then the user
 ought to be able to adjust the sort criteria without any
@@
 
-This is in stark contrast to the status quo, where each
-system targets one stage only.  The Website Meta Language
-(WML) \cite{engelschall-wml} is an ``off-line HTML
-generation toolkit'' designed for the \emph{publish} stage.
-But many other programs (and countless ad-hoc scripts) spit
-out HTML pages: think of \LaTeX2HTML.  Google
-reports surprisingly many
+This approach is in stark contrast to the status quo, where each
+system targets one stage only.  One such system, the Website Meta
+Language~\cite{engelschall-wml}, is an ``off-line HTML generation
+toolkit'' designed for the \emph{publish} stage.  But many other
+programs (and countless ad-hoc scripts) spit out HTML pages:
+\LaTeX2HTML, for example.  Google reports surprisingly many
 programs\footnote{\url{http://google.com/search?q=gedcom+generate+html}}
-for creating family tree web sites from genealogy database
-files; these count as publish-stage tools.
+for creating family tree web sites from genealogy database files;
+these also count as publish-stage tools.
 
@@
 of the overhead of CGI.
-%%
+
 There is, relatively, a paucity of languages that operate
@@
 Imagine implementing the conference calendar, as conceived
-above, using currently deployed technology: A Perl script
+above, using currently deployed technology: a Perl script
 outputs a PHP page which contains JavaScript!  Values are
@@
 
-We present ``\MOC server pages,'' a new domain-specific
-language for web applications programming.  It leverages the
-staging annotations of \MOC~\cite{calcagno03meta,taha00metaml}
-to provide safe and precise control over the each stage of the
-computation.  We have implemented the system in two parts: our
-translator transforms the server page language into a \MOC
-module, which then can be incorporated into our multi-threaded
-HTTP 1.1 server (also written in \MOC).  The scalability
-gained by staging certain applications is stunning.
+We present ``\MOC server pages,'' a new domain-specific language for
+web applications programming.  It leverages the staging annotations of
+\MOC~\cite{calcagno03meta,taha00metaml} to provide safe and precise
+control over the first two stages.  (We leave further consideration of
+the display stage as future work.)  The system is implemented as two
+components: a translator transforms the server page language into a
+\MOC module, which then can be incorporated into our multi-threaded
+HTTP/1.1 server (also written in \MOC).  The scalability gained by
+staging certain applications is stunning.
 
@@
 object which is subsequently run on each request (serve stage).
+The third (display) stage proposed in section~\ref{sec:intro} is not
+supported by this design.
 Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
@@
 
-\mylst{trans.ml}{This is the translation of the
-    page in figure~\ref{lst:trans.meta}.  Declarations have
-    been lifted, and everything else appears within
-    \cd{\Mbegincode brackets\Mendcode}}
+\mylst{trans.ml}{An automatic translation of the page in
+  listing~\ref{lst:trans.meta}.  Declarations have been lifted, and
+  everything else appears within \cd{\Mbegincode brackets\Mendcode}}
 
@@
 contained there are executed when the page is published.  The
-constructed code begins with the \cd{{\KWfont let} page} line in
-listing~\ref{lst:trans.ml}.
+constructed code follows the line that begins with \cd{{\KWfont let} page}.
 
@@
 details, including the headers and query arguments.  \cd{puts} is a
-function, provided by the server, to 
-transmit text across the network to the user's browser.  \cd{arg} is
-then defined to look up an argument in the query string of the
-request; its type is \cd{string\to string option}.
-
-These are all \emph{serve} (second)
-stage values.  It makes no sense to access them in the
-\emph{publish} (first) stage, and indeed the \MOC type system
-prevents this.
+function, provided by the server, to transmit text across the network
+to the user's browser.  \cd{arg} is then defined to look up an
+argument in the query string of the request; its type is \cd{string\to
+  string option}.
+Since \cd{req} and \cd{puts} are serve-stage arguments, it is
+incorrect to invoke them in the \emph{publish} (first) stage, and indeed
+the \MOC type system prevents this.
 
@@
 listing~\ref{lst:trans.ml} would be dotted with %
-\cd{puts}~\verb+"\n"+ statements).  It automatically follows
-code blocks with either a semi-colon or the {\KWfont in}
-keyword, as required.
+\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
+semi-colon or the {\KWfont in} keyword to code blocks, as required (see,
+for example, the lines containing \cd{statements},
+\cd{more\_statements}, and \cd{{\KWfont let} x} in
+listing~\ref{lst:trans.ml}).
 
@@
 Now, all that remains to execute at serve time is %
-\cd{puts "0755"}.  Finally it is possible to splice in serve-stage
+\cd{puts "0755"}.  It is also possible to splice in serve-stage
 parameters at publish time, such as in this count-down example:%
@@
 \subsection{Publish-stage arguments}
-One last feature in the design of \MOC server pages is the ability
-to take publish-stage arguments.  The parameter names are specified to
-the translator in a \emph{pragma} declaration, which can appear
-anywhere in the code:
 
-(EXPLAIN THIS INSTEAD USING TRANS FIGURE)
+\MOC server page support publish-stage arguments, as demonstrated by
+the identifiers \cd{a b c} in listings~\ref{lst:trans.meta}
+and~\ref{lst:trans.ml}.  The programmer specifies the pattern to be
+used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
+may appear anywhere in the code.
 
-Now the code can be instantiated with early parameters in countless
-ways.  For example, requesting the URI \id{/longcount} could print a
-count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
-are generated from the same server page.  Mapping from URIs to
-instantiated \MOC code is, for now, an implementation detail
-(see section~\ref{sec:impl}).
+With publish-stage arguments, the page may be instantiated in
+countless ways.  Continuing the count-down example, we could map the
+URI \id{/longcount} to a code block generated with an actual argument
+of 99, while \id{/shortcount} refers to code with the actual
+argument 9.  Both are generated from the same server page.
+
+\lgrindfile{count2.meta}
+
+Mapping from URIs to instantiated \MOC code is, for now, left as an
+implementation detail (see section~\ref{sec:impl}).
 
@@
 page title includes a dynamic argument, then \cd{preamble} will have
-to be delayed until theh serve stage.
+to be delayed until the serve stage.
 
@@
 
-\mylst{power.meta}{This implements the staged power function as a
-  web service.  The exponent \cd{y} is an early (publish stage)
-  argument; but the base \cd{x} is late (serve stage).}
-
 Judging from its prevalence in the literature, we are certain that
@@
 selecting the exponent, the user types the base into the form and
-presses return.  The result is computed using OCaml's
-arbitrary-precision \cd{Num} module.  The complete, documented script
-appears in listing~\ref{lst:power.meta}.
+presses return.  The result is computed using the arbitrary-precision
+\cd{Num} module of the OCaml library.  The complete script appears in
+listing~\ref{lst:power.meta}.
+
+\mylst{power.meta}{This implements the staged power function as a
+  web service.  The exponent \cd{y} is an early (publish stage)
+  argument; but the base \cd{x} is late (serve stage).}
 
 The user's input is converted to a number relatively late in the
-script.  If \cd{num\_of\_string} generates an exception (perhaps
-because the user typed non-numeric text into the box), the navigation
-bar and form will have already been output before the error message
-appears.
+script; see the line marked with {(****)}.  If \cd{num\_of\_string}
+generates an exception (perhaps because the user typed non-numeric
+text into the box), the navigation bar and form will have already been
+output before the error message appears.
 
@@
 The \cd{list} is formed by testing filenames (in reverse alphabetical
-order) against the compiled regular expression \cd{rc}, including only
+order) against the compiled regular expression \cd{rc}, prepending only
 those that match.  The file info record for each file has already
@@
 
-Currently, the \MOC implementation is not capable of reading
-and writing \cd{code} values to a file.  This means that
-generated server pages (after the publish stage) must already be
-in the server memory when it starts.  This is an unfortunate
-limitation, as it means we must restart the entire web server to
-re-publish any particular page!
+The implementation consists of two parts: a translator and a web
+server.  The translator, implemented in Standard ML,\footnote{Mainly
+  because the author has considerable experience with the ML-Yacc that
+  comes with SML/NJ, and is reluctant to learn the OCaml version!}
+transforms the server page syntax into plain \MOC, as illustrated in
+listing~\ref{lst:trans.ml}.  It recognizes all the server page blocks
+defined in section~\ref{sec:design} but does not completely parse the
+\MOC code contained within them.  Instead, it recognizes just enough
+of the keywords and delimiters to decide whether to add semi-colons or
+`{\KWfont in}' after each block.  This approach makes the translator
+fairly robust to minor changes in OCaml syntax.
 
-A generated server page has this type:
+The disadvantage of this method is that very few syntax errors (and
+no type errors) are detected by the translator itself.  So, errors
+reported by \MOC are displayed in terms of the translated program, not
+the source program.
+
+The server component is much more complex.  We implemented the
+essential parts of the HTTP/1.1 specification as a multi-threaded
+OCaml program.  Upon receiving a request for some URI, the server uses
+a \emph{chain of responsibility}~\cite{gamma95design} to determine
+how to handle it.  The chain is specified as a parameter to the
+server.  Two primary handlers are provided: \cd{FileHandler} and
+\cd{CodeHandler}.
+
+The \cd{FileHandler} takes a file system path as a parameter, and then
+interprets URIs as file names relative to that path.  If such a file
+exists, it sends it verbatim to the client.  If not, it passes
+the responsibility on to the next handler in the chain.
+
+The \cd{CodeHandler} is provided with a dictionary to map URIs to code
+values.  It looks up the URI in the dictionary, and if a match is
+found it invokes that code.  For now, the map is hard-coded at build
+time and the code values are already in memory when the server starts
+accepting requests.  This is because \MOC is currently not capable of
+reading and writing code values to a file.
+
+Here are the types of the page code and the map, along with signatures
+for the \cd{CodeHandler} (and two variants):
 
 \lgrindfile{codeHandler.mli}
+%%
+\cd{page} is the type of the code that is constructed by each server
+page; recall that the page takes two serve-time arguments: one
+encapsulating the HTTP request, the other is the \cd{puts} function
+for sending text back to the client.
 
-As the server starts, it constructs a dictionary (\cd{code
-  StringMap.t}) mapping URIs to code.  This dictionary is
-consulted on each HTTP request.  If the URI is not found there,
-then we check the file system, in case it refers to a static
-file.  If this fails, we respond with a 404 (resource not
-found).
+The \cd{map} type maps a string (the URI) to a pair containing a
+function to (re-)generate the page (by running the first stage) and a
+ref cell where the page is cached.  No \cd{code} types show up here,
+but they are used within the \cd{unit\to page} function.  This
+function is implemented using the \Mrun{} (run) operator.  Passing
+around code values themselves is difficult because they must remain
+polymorphic to run them, and OCaml does not support rank-2
+polymorphism.
 
@@
 Finally, we look at the performance of staged and unstaged directory
-browsing; see figure~\ref{fig:browse}.  For this task, we created
-directories containing a fixed number of files with random data.  The
-average file size was 32k.  The \emph{x}~axis shows the number of
-files in the directory.  Besides the staging factor, two
-implementation languages are compared: we coded the same functionality
-in PHP.  The `staged PHP' version must be run first from the command
-line; this outputs another PHP script which is then run by the server.
+browsing; see figure~\ref{fig:browse}.  Here, we created directories
+containing fixed numbers of files with random data.  The average file
+size was 32k.  The \emph{x}~axis shows the number of files in each
+directory.
+
+Besides the staging factor, two implementation languages are compared:
+we implemented the same functionality in PHP.  The `staged PHP' version must
+be run first from the command line; this outputs a PHP script
+which is then run by the server.  With PHP we must, as noted in
+section~\ref{sec:intro}, manage the quoting and persistence issues
+by hand.  The staged PHP code contains gems like this:
+\begin{verbatim}
+<?= "<?\n" ?>
+<?= "\$list = unserialize(\"".
+    addcslashes(serialize($list),'"').
+    "\");\n" ?>
+<?= "?>\n" ?>
+\end{verbatim}
+where the serialize occurs in stage one and the unserialize in stage two.
 
@@
 of files in the directory.  With \MOC, the code size is also
-proportional to the number of files, but the values are already in
-memory.
+proportional to the number of files, but the values are byte-compiled
+(and already in memory).  At any rate, our goal is not to beat PHP on
+performance, but rather to gain the performance benefit of staging
+\emph{without} the awkwardness of staging in a language (like PHP) that
+does not support it.
 
@@
 
-Web publishing is an important application domain that is
-naturally staged.  Web programmers write staged programs every
-day, but they do it the old-fashioned way: one script outputs
-another as a string.  This is a tremendous opportunity for
-multi-stage languages.
+Web publishing is an important application domain that is naturally
+staged.  Web programmers write staged programs, but they do it the
+old-fashioned way: one script outputs another as a string.  This is a
+tremendous opportunity for multi-stage languages.
 

### 2007-04-16 paper/paper.tex
svn-r28: 1st journal revision
darcs-hash:20070416223336-e5a07-01112a05bd1eb888105e8ace6bcd3740ef742de7.gz

@@
-\documentclass[preprint]{acm_proc_article-sp}
-\usepackage{ucs,textcomp,url,xspace,graphicx,color}
-\usepackage[nolineno]{lgrind}
-\usepackage[bf]{caption2}
+\documentclass{elsart}
+\usepackage{textcomp,url,xspace,graphicx,color}
+\def\UrlFont{\tt\small}
+\usepackage[utf8]{inputenc}
+\usepackage[leftno,norules,nolineno]{lgrind}
+%\usepackage[bf]{caption2}
 \usepackage[numbers]{natbib}
-\usepackage[utf8]{inputenc}
 \ifx\pdfoutput\undefined\else
@@
   {\centering%
-    \includegraphics[width=20pc,keepaspectratio]{#1}}}
+    \includegraphics[width=25pc,keepaspectratio]{#1}}}
 \newcommand{\gnuplot}[2]{%
-  \myfig{#1}{#2}{\input{#1}}}
-\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}
+  \myfig{#1}{#2}{\centering\input{#1}}}
 
-\newcounter{listing}
-\setcounter{listing}{0}
-\makeatletter
-\def\ext@listing{lol}
-\def\fnum@listing{Listing\nobreakspace\thelisting}
-\makeatother
+\LGindent=16pt
+\def\lgrindfile#1{
+  \par\vskip.5\baselineskip\begingroup
+  \begin{lgrind}\input #1\relax
+  \end{lgrind}\endgroup}
+
+\newcommand{\mylstf}[2]{%
+  \myfig{#1}{\texttt{#1}.  #2}{\LGnuminterval=3\lgrindfile{#1}}}
 \newcommand{\mylst}[2]{%
-  \captionof{listing}{\textbf{#1.} #2}%
-  \label{lst:#1}%
-  \lgrindfile{#1}}
-\newcommand{\mylstf}[2]{%
-  \begin{figure}[tbp!]%
-    \lgrindfile{#1}%
-    \captionof{listing}{\textbf{#1.} #2}%
-    \label{lst:#1}%
-  \end{figure}}
-
-\title{\MOC Server Pages:}
-\subtitle{Web Publishing as Staged Computation}
-\numberofauthors{1}
-\author{
-  \alignauthor Christopher League\\
-  \affaddr{Long Island University · Computer Science}\\
-  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
-  \email{christopher.league@liu.edu}}
-%\conferenceinfo{\MOC Workshop}{October 2004, Vancouver}
-%\CopyrightYear{2004}
-%\toappear{Submitted to the \MOC Workshop (Vancouver,
-%  2004).}
+  \myfig{#1}{#2}{\LGnuminterval=3\lgrindfile{#1}}}
 
 \begin{document}
-\maketitle
+\begin{frontmatter}
+\title{\MOC Server Pages:\\Web Publishing as Staged Computation}
+\author{Christopher League}
+\address{Long Island University \textperiodcentered{} Computer Science\\
+  1 University Plaza \textperiodcentered{} Brooklyn, NY 11201\\
+  \textup{\texttt{christopher.league@liu.edu}}}
+
 \begin{abstract}
-  Modern dynamic web services are really computer programs.
-  Some parts of these programs run off-line, others run
-  server-side on each request, and still others run within
-  the browser.  In other words, web publishing is
-  \emph{staged computation,} either for better performance,
-  or because certain resources are available in one stage
-  but not another.  Unfortunately, the various web
-  programming languages target one stage only.  This is a
-  tremendous opportunity for multi-stage languages in
-  general, and for \MOC in particular.
+  Modern dynamic web services are really computer programs.  Some
+  parts of these programs run off-line, others run server-side on each
+  request, and still others run within the browser.  In other words,
+  web publishing is \emph{staged computation,} either for better
+  performance, or because certain resources are available in one stage
+  but not another.  Unfortunately, the various web programming
+  languages make it difficult to spread computation over more than one
+  stage.  This is a tremendous opportunity for multi-stage languages
+  in general, and for \MOC in particular.
 
-  We present the design of \MOC Server Pages.  Unlike
-  other languages in its genre, the embedded \MOC code
-  blocks may be annotated with staging information, so that
-  the programmer may safely and precisely control which
-  computation occurs in which stage.  A prototype web
-  server, written in OCaml, supports web sites with both
-  static and dynamic content.  We provide several sample
-  programs and demonstrate the performance gains won using
-  multi-stage programming.
+  We present the design of \MOC Server Pages.  Unlike other languages
+  in its genre, the embedded \MOC code blocks may be annotated with
+  staging information, so that the programmer may safely and precisely
+  control which computation occurs in which stage.  A prototype web
+  server, written in OCaml, supports web sites with both static and
+  dynamic content.  We provide several sample programs and demonstrate
+  the performance gains won using multi-stage programming.
 \end{abstract}
-\category{D.3.2}{Programming Languages}{Language
-  Classifications}[Specialized application languages]
-\category{H.3.5}{Information Storage and Retrieval}{Online Information
-  Services}[Web-based services]
+\end{frontmatter}
+
+%\category{D.3.2}{Programming Languages}{Language
+%  Classifications}[Specialized application languages]
+%\category{H.3.5}{Information Storage and Retrieval}{Online Information
+%  Services}[Web-based services]
 
@@
 on the web: \emph{performance matters.}  As a web publisher, visitors
-are your bread and butter.  But will your servers and scripts be ready
-for the day that your site is featured on prime time television, or on
+are your livelihood.  But will your servers and scripts be ready for
+the day that your site is featured on prime time television, or on
 \texttt{slashdot.org}?  If tens of thousands of potential users drop
@@
 bandwidth limitations, Slashdot creator Rob Malda says most
-sites that fall suffer from poor planning and architecture:
+sites that fail suffer from poor planning and architecture:
 \begin{quote}
   The problem is, if you have any sort of complicated code
-  on your page, that’s what kills people. I generally think
+  on your page, that's what kills people. I generally think
   that sites are not dying because they fill their pipe, but
@@
   As a general rule, anybody who has a pretty good
-  understanding of web design, they’ve done a good job of
+  understanding of web design, they've done a good job of
   learning what information to cache. What information needs
-  to be pregenerated. So when you’re actually loading a
-  page, even if it’s a complicated page that looks dynamic
-  and custom, on the back end of that, what they’re really
+  to be pregenerated. So when you're actually loading a
+  page, even if it's a complicated page that looks dynamic
+  and custom, on the back end of that, what they're really
   doing is putting together a bunch of puzzle pieces that
@@
 
-This quote points (informally) to the idea of \emph{staging} the
-computation performed by a web service.  Indeed, web publishing is an
-application area that is naturally staged:\footnote{Nevertheless, it
-  seems the literature on meta-programming has not yet acknowledged
-  web services as a potential application area, although
-  \citet{sheard01accomp} does
-  mention \emph{mobile code.}} %
+Malda's observation points (informally) to the idea of \emph{staging}
+the computation performed by a web service.  Indeed, web publishing is
+an application area that is naturally staged:\footnote{Nevertheless,
+  it seems the literature on meta-programming has not yet acknowledged
+  web services~\cite{w3c-wsdl} as a potential application area,
+  although \citet{sheard01accomp} does mention \emph{mobile code.}} %
 \begin{enumerate}
@@
 
-\screenshot{confcal}{A simple conference calendar, implemented as a
-  PHP script that connects to a PostgreSQL database.  Not as
-  sophisticated as the service described in the text.}
+\screenshot{confcal}{A simple conference calendar web service.}
+%, implemented as a
+%  PHP script that connects to a PostgreSQL database.  It is not as
+%  sophisticated as the service described in the text.}
 
@@
 presents some identification (in the form of a `cookie'),
-but since the conference data changes infrequently, it may
+but since the conference data change infrequently, it may
 help to prepare the text of each row in advance.
@@
 
-At this point, we should emphasize the importance of
-\emph{profiling} in developing scalable web services.  This
-particular design for the conference calendar may not be
-optimal, depending on the relative speeds of the CPU,
-memory, database, network, and disk.  Rather, our aim is to
-provide a single language in which the various staging
-possibilities can be expressed naturally.
+At this point, we should emphasize the importance of \emph{profiling}
+in developing scalable web services.  This particular design for the
+conference calendar may not be optimal, depending on the number of
+entries and the relative speeds of the CPU, memory, database, network,
+and disk.  Rather, our aim is to provide a single language in which
+the various staging possibilities can be expressed naturally.
 
 This approach is in stark contrast to the status quo, where each
-system targets one stage only.  One such system, the Website Meta
-Language~\cite{engelschall-wml}, is an ``off-line HTML generation
-toolkit'' designed for the \emph{publish} stage.  But many other
-programs (and countless ad-hoc scripts) spit out HTML pages:
+system targets one stage only.  The Website Meta
+Language\footnote{\url{http://www.thewml.org/}} is an ``off-line HTML
+generation toolkit'' designed for the \emph{publish} stage.  But many
+other programs (and countless ad-hoc scripts) spit out HTML pages:
 \LaTeX2HTML, for example.  Google reports surprisingly many
@@
 The \emph{serve} stage is well-served by the ``server page''
-languages, including JSP~\cite{sun-jsp},
-ASP~\cite{microsoft-asp},
-MSP~\cite{elsman03web}, and
-PHP~\cite{bakken04php}.  The CGI protocol~\cite{w3c-cgi}
-addresses the \emph{serve} stage, as do the embedded
-interpreters (such as \texttt{mod\_perl} and
-\texttt{mod\_python} for Apache) that exist to ameliorate some
-of the overhead of CGI.
+languages, including
+JSP,\footnote{\url{http://java.sun.com/products/jsp/}}
+ASP,\footnote{\url{http://www.asp.net/}} and
+PHP.\footnote{\url{http://www.php.net/}} The Common Gateway Interface
+(CGI)\footnote{\url{http://www.w3.org/CGI/}} addresses the
+\emph{serve} stage, as do the embedded interpreters (such as
+\texttt{mod\_perl} and \texttt{mod\_python} for Apache) that exist to
+ameliorate some of the overhead of CGI.
 
-There is, relatively, a paucity of languages that operate
-client-side (\emph{display} stage), probably due to the
-difficulty of securing an installed base of interpreters.
-JavaScript~\cite{netscape-javascript} and Java applets are
+There is, relatively, a paucity of languages that operate client-side
+(\emph{display} stage), probably due to the difficulty of securing an
+installed base of interpreters.  JavaScript and Java applets are
 notable exceptions.
@@
 above, using currently deployed technology: a Perl script
-outputs a PHP page which contains JavaScript!  Values are
+outputs a PHP page which embeds JavaScript!  Values are
 passed from one stage to the next as strings, and the
@@
 
+Strictly speaking, these languages are not exclusively confined to the
+stages that I have indicated: Javascript can be run client-side, PHP
+can be run off-line, and so on.  Nevertheless, migrating code between
+stages is hard, and the need for quoting and persistence are
+practically show-stoppers.  For comparison purposes (further described
+in section~\ref{sec:perf}), we staged some code using PHP.  To achieve
+persistence of composite variables from one stage to the next, it
+contains gems like this:\label{php-gem}
+\begin{verbatim}
+    <?= "<? \$list = unserialize(\"".
+        addcslashes(serialize($list),'"').
+        "\"); ?>\n" ?>
+\end{verbatim}
+where the \texttt{serialize} library function occurs in stage one and
+the \texttt{unserialize} in stage two.  Notice that the \verb+$+
+preceding the first occurrence of \verb+list+ is quoted, but the
+second occurrence is not.  The \texttt{addcslashes} function is needed
+in case the serialized representation contains special characters
+(such as double quotes or newlines) that would be misinterpreted by
+the PHP parser in the next stage.  Interpreting the stage-one program
+guarantees nothing about the well-formedness of the stage-two program
+(generated as a string).
+
 We present ``\MOC server pages,'' a new domain-specific language for
-web applications programming.  It leverages the staging annotations of
-\MOC~\cite{calcagno03meta,taha00metaml} to provide safe and precise
-control over the first two stages.  (We leave further consideration of
-the display stage as future work.)  The system is implemented as two
-components: a translator transforms the server page language into a
-\MOC module, which then can be incorporated into our multi-threaded
-HTTP/1.1 server (also written in \MOC).  The scalability gained by
-staging certain applications is stunning.
+web applications programming.  It leverages the staging annotations
+and static typing of \MOC~\cite{calcagno03meta,taha00metaml} to
+provide safe and precise control over the first two stages.  (We leave
+further consideration of the display stage as future work.)  The
+system is implemented as two components: a translator transforms the
+server page language into a \MOC module, which then can be
+incorporated into our multi-threaded HTTP/1.1 server (also written in
+\MOC).  The scalability gained by staging certain applications is
+stunning: in section~\ref{sec:perf} we describe a directory browsing
+service where staging yields a factor of 30 improvement in throughput.
+The unstaged version would certainly succumb to the Slashdot effect.
 
@@
 section~\ref{sec:impl}.  Performance and scalability are
-discussed in section~\ref{sec:perf}
+discussed in section~\ref{sec:perf}.
 
@@
 PHP programmers are familiar with this syntax for embedding code, but
-in our case, the code itself is written in \MOC.  Here is your first
+in our case, the code itself is written in \MOC.  Here is a trivial
 \MOC server page:
@@
 \end{quote}
-If a code block raises an exception, the message is sent to the
-client's browser in bold-face, and the rest of the page is
-aborted.
+The OCaml function \cd{failwith} raises a \cd{Failure} exception
+containing the provided message.  If a code block raises an exception,
+the message is sent to the client's browser in bold-face, and the rest
+of the page is aborted.
 
-A very common use of code blocks is to print out (i.e., send to
-the browser) the result of evaluating some expression.  The
-syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
-have type \cd{string}.  Alternatively, messages may be formatted
-with \cd{printf} by placing the format string immediately after
-the code delimiter: \lgrindfile{pi.meta}
+A very common use of code blocks is to print out (i.e., send to the
+browser) the result of evaluating some expression.  The syntax
+\cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must have
+type \cd{string}.  Alternatively, messages may be formatted with
+\cd{printf} by placing the format string immediately after the code
+delimiter.\footnote{In a departure from the standard OCaml syntax for
+  \cd{printf}, the arguments are comma-delimited.  This way, fewer
+  parentheses are required when using the escape or lift operators on
+  the arguments.}
+%%
+\lgrindfile{pi.meta}
+The output is:
 \begin{quote}
-  ........ $\pi$ ÷ 008 is 0.3927
+  ........ $\pi$ $\div$ 008 is 0.3927
 \end{quote}
-In a departure from the standard OCaml syntax for \cd{printf}, the
-arguments are comma-delimited.  This way, fewer parentheses are
-required when using the escape or lift operators on the arguments.
 
@@
 \lgrindfile{perm.meta}
+Output:
 \begin{quote}
@@
 
-Once published, this output will never change!  The \cd{stat}
-call is executed only once, not on each request.
+Once published, this output will never change!  The \cd{stat} call is
+executed only once (because it is in a declaration block), not on each
+request.  This is already a rudimentary kind of staging, but with the
+annotations of \MOC, we will gain both flexibility and safety, as
+we'll see in the rest of this section.
 
@@
 performed immediately (because it is escaped), and the result is
-spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}.
+spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}
 
@@
 \subsection{Translation to \MOC}
-To see how all this works, let's consider how a \MOC server
+To see how all this works, consider how a \MOC server
 page is translated into a proper \MOC program, to be executed
@@
 The third (display) stage proposed in section~\ref{sec:intro} is not
-supported by this design.
-Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
-and listing~\ref{lst:trans.ml} contains its translation.
+yet supported by this design.
+Figure~\ref{fig:trans.meta} shows a sample \MOC server page,
+and figure~\ref{fig:trans.ml} contains its translation.
 
-\mylst{trans.meta}{This file demonstrates the various
+\mylstf{trans.meta}{This file demonstrates the various
   kinds of code blocks.}
 
-\mylst{trans.ml}{An automatic translation of the page in
-  listing~\ref{lst:trans.meta}.  Declarations have been lifted, and
-  everything else appears within \cd{\Mbegincode brackets\Mendcode}}
+\mylstf{trans.ml}{An automatic translation of the page in
+  figure~\ref{fig:trans.meta}.}
+%  Declarations have been lifted, and
+%  everything else appears within \cd{\Mbegincode brackets\Mendcode}}
 
@@
 contained there are executed when the page is published.  The
-constructed code follows the line that begins with \cd{{\KWfont let} page}.
+constructed code begins on line~5.
 
@@
 function, provided by the server, to transmit text across the network
-to the user's browser.  \cd{arg} is then defined to look up an
-argument in the query string of the request; its type is \cd{string\to
-  string option}.
-Since \cd{req} and \cd{puts} are serve-stage arguments, it is
-incorrect to invoke them in the \emph{publish} (first) stage, and indeed
-the \MOC type system prevents this.
+to the user's browser.  
+%%
+The library function \cd{Request.arg} of the type \cd{request\to
+  string\to string option} looks up the string value of the request
+parameter with a given name.  Since \cd{req} and \cd{puts} are
+serve-stage arguments, it is incorrect to use them in the
+\emph{publish} (first) stage, and indeed the \MOC type system prevents
+this.
 
@@
 immediately follow code blocks (otherwise,
-listing~\ref{lst:trans.ml} would be dotted with %
+figure~\ref{fig:trans.ml} would be dotted with %
 \cd{puts}~\verb+"\n"+ statements).  It automatically appends a
-semi-colon or the {\KWfont in} keyword to code blocks, as required (see,
-for example, the lines containing \cd{statements},
-\cd{more\_statements}, and \cd{{\KWfont let} x} in
-listing~\ref{lst:trans.ml}).
+semi-colon or the {\KWfont in} keyword to code blocks, as required
+(see, for example, lines 7, 11, and 12 in figure~\ref{fig:trans.ml}.)
+%%
+Finally, because the translator partially parses the OCaml code
+blocks, it is not confused by code delimiters within strings and
+comments, nor by other uses of \la{} and \ra{} as operators.
 
@@
 program, the effects of adding \MOC staging operators to our pages
-should be predictable.  Below is a reformulation of the permissions
-example, using escape \Msplice{} and \cd{lift} to ensure that
-\cd{stat} is called only once.%
-\lgrindfile{perm2.meta}%
+should be predictable.  Below is another example using \cd{stat}, this
+time to display the size of the server log file.
+With the serve-stage argument \texttt{unit}, we can specify whether
+the size should be expressed in bytes (the default), kilobytes, etc.
+\lgrindfile{size.meta}%
+
+To stage this, the \cd{stat} and \cd{float\_of\_int} calls could be
+lifted into the declaration block, as with the permissions example.
+Furthermore, we may also use \cd{lift} and the splice operator to
+perform the divisions in advance, even though they are underneath the
+{\KWfont match} (which cannot happen until the serve stage):
+\lgrindfile{size2.meta}%
+
 Still, the \cd{printf} conversion is performed at serve time, so maybe
 it is best to pre-generate that as well:%
-\lgrindfile{perm3.meta}%
-Now, all that remains to execute at serve time is %
-\cd{puts "0755"}.  It is also possible to splice in serve-stage
-parameters at publish time, such as in this count-down example:%
+\lgrindfile{size3.meta}%
+
+Now, all that remains to execute at serve time is to check the dynamic
+\cd{unit} argument and spit out one of three pre-generated strings.
+The generated code block looks something like this (substantially
+cleaned up from the \MOC pretty-printer, with references to persistent
+values resolved in-place):
+\lgrindfile{size-out.ml}%
+
+To handle more complex situations, code blocks may be constructed
+conditionally and recursively.  Here is an example that
+prints a count-down, but removes the loop overhead by expanding to a
+sequence of 99 \cd{puts} statements.
 \lgrindfile{count.meta}%
-It removes the loop overhead by expanding to a sequence of 99
-\cd{puts} statements.
 
-These tiny examples illustrate some of the common uses of the
-escape operator, for which we developed syntactic sugar.  The first
-set of code blocks we define use the tilde character to indicate that
-some part of the enclosed quote is escaped:
+The brackets around the \cd{puts} on the last line are necessary,
+because \cd{count} splices the \cd{puts} call into the code it
+generates.  Omitting the brackets would result in a compile-time type
+error.
+
+These tiny examples suggest the ease with which the `boundary' between
+the stages can be adjusted, just by tweaking the staging annotations.
+Moreover, the static typing of \MOC ensures in advance that our
+programs generate type-correct code only.  
+%%
+The examples also illustrate some of the common uses of the escape
+operator, for which we developed syntactic sugar.  The first set of
+code blocks we define use the tilde character to indicate that some
+part of the enclosed code is escaped:
 
@@
 \\
-\cd{\Mquote{\Mtilde}let x = c \Munquote} &
-\cd{\Mquote{} let x = \Msplice( c ) \Munquote}
+\cd{\Mquote{\Mtilde}{\KWfont let} x = c \Munquote} &
+\cd{\Mquote{} {\KWfont let} x = \Msplice( c ) \Munquote}
 \\
@@
 \\
-\cd{\Mquote{\Mbang}let x = c \Munquote} &
-\cd{\Mquote{} let x = \Msplice(lift( c )) \Munquote}
+\cd{\Mquote{\Mbang}{\KWfont let} x = c \Munquote} &
+\cd{\Mquote{} {\KWfont let} x = \Msplice(lift( c )) \Munquote}
 \\
@@
 
-We will see examples of most of these blocks in section~\ref{sec:eg}.
+We will see examples of most of these blocks in section~\ref{sec:eg}
+and appendix~\ref{sec:dir.meta}.
 
@@
 
-\MOC server page support publish-stage arguments, as demonstrated by
-the identifiers \cd{a b c} in listings~\ref{lst:trans.meta}
-and~\ref{lst:trans.ml}.  The programmer specifies the pattern to be
+\MOC server pages support publish-stage arguments, as demonstrated by
+the identifiers \cd{a b c} in figures~\ref{fig:trans.meta}
+and~\ref{fig:trans.ml}.  The programmer specifies the pattern to be
 used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
@@
 
-To explore the expressiveness of this design, we now look at a
-series of web services, organized to comprise a small web
-site.\footnote{A demo is available at
-  \url{http://comsci.liu.edu/metaserv/}}
-To give the services a similar look, we developed a site-wide style
-sheet, and an OCaml function to generate a navigation bar from a
-hierarchical list of page titles and links.  The following bindings are
-available site-wide:
-
-\lgrindfile{navspec-sig.ml}
+To explore the expressiveness of this design, we now look at a series
+of web services, organized to comprise a small web site.\footnote{A
+  demo is available at \url{http://comsci.liu.edu/metaserv/}} To give
+the services a similar look, we developed a site-wide style sheet, and
+an OCaml function to generate a navigation bar from a hierarchical
+list of page titles and links.  
 %%
-The functions above may be invoked at any stage.  Naturally, if the
-page title includes a dynamic argument, then \cd{preamble} will have
-to be delayed until the serve stage.
+\mylst{navspec-sig.ml}{These functions help define the standard site
+  layout.}
+%%
+The page layout functions in figure~\ref{fig:navspec-sig.ml} may be
+invoked at any stage.  Naturally, if the page title includes a dynamic
+argument, then \cd{preamble} will have to be delayed until the serve
+stage.
 
@@
 \screenshot{power127}{Web browser displaying the result of
-  7\textsuperscript{127}.  The URI \texttt{/power127} maps to code
-  where the early argument (the exponent) has been set to 127, and the
-  late argument (\texttt{x}, the base) is taken from the query
-  string.}
+  7\textsuperscript{127}.}
+%  The URI \texttt{/power127} maps to code
+%  where the early argument (the exponent) has been set to 127, and the
+%  late argument (\texttt{x}, the base) is taken from the query
+%  string.}
 
-Judging from its prevalence in the literature, we are certain that
-millions of grateful users would subscribe to an online service
-capable of computing the \emph{power} function!  The screen shot in
-figure~\ref{fig:power127} illustrates how it works.  The navigation
-bar shows the exponents for which code has been
+Judging from its prevalence in the multi-stage programming literature,
+we are certain that millions of grateful users would subscribe to an
+online service capable of computing the \emph{power} function.  The
+screen shot in figure~\ref{fig:power127} illustrates how it works.
+The navigation bar shows the exponents for which code has been
 pre-generated.\footnote{In the current implementation, it is not
@@
 \cd{Num} module of the OCaml library.  The complete script appears in
-listing~\ref{lst:power.meta}.
+figure~\ref{fig:power.meta}.
 
-\mylst{power.meta}{This implements the staged power function as a
-  web service.  The exponent \cd{y} is an early (publish stage)
-  argument; but the base \cd{x} is late (serve stage).}
+\mylstf{power.meta}{The staged power function
+ as illustrated in figure~\ref{fig:power127}.}
+%  web service.  The exponent \cd{y} is an early (publish stage)
+%  argument; but the base \cd{x} is late (serve stage).}
 
 The user's input is converted to a number relatively late in the
-script; see the line marked with {(****)}.  If \cd{num\_of\_string}
+script (line 21).  If \cd{num\_of\_string}
 generates an exception (perhaps because the user typed non-numeric
@@
 \cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
-brackets and escapes from the code in listing~\ref{lst:power.meta}.
+brackets and escapes from the code in figure~\ref{fig:power.meta}.
 
@@
 
-This kind of service can be fairly intense; one HTTP request is likely
-to generate dozens of system calls and disk accesses.  If the
-directory is viewed more often that it is changed, then it makes sense
-to cache or pre-generate the pages.  Although it is simple enough to
-write a script to generate directory pages off-line, what if we also
-want dynamic behavior, such as user-controlled sorting and filtering?
+This kind of service can be fairly resource intensive; each HTTP
+request is likely to generate dozens of system calls and disk
+accesses.  If the directory is viewed more often than it is changed,
+then it makes sense to cache or pre-generate the pages.  Although it
+is simple enough to write a script to generate static directory pages
+off-line, what if we also want dynamic behavior, such as
+user-controlled sorting and filtering?
 
@@
 
-\screenshot{server-dir}{Browsing the server source code directory.
-  File names are colored according to their \emph{kind,} which is
-  computed from the extension and the file type.  See also
-  listing~\ref{lst:dir.meta}.}
+\screenshot{server-dir}{Browsing the server source code directory.}
+%  File names are colored according to their \emph{kind,} which is
+%  computed from the extension and the file type.  See also
+%  figure~\ref{fig:dir.meta}.}
 
@@
 of each remaining row.  The files need not be opened or even
-\id{stat(2)}ed while the user is waiting.  Listing~\ref{lst:dir.meta}
+\id{stat(2)}ed while the user is waiting.  Appendix~\ref{sec:dir.meta}
 contains the complete script, with documentation.
 
-\mylst{dir.meta}{This implements the directory browser shown in
-  figure~\ref{fig:server-dir}.}
+Much of the code is unaffected by staging; helper functions (that sort
+according to the selected criterion, format the human-readable sizes,
+and collect the file information) are oblivious to the stage in which
+they are run.  Therefore, the staged code comprises a relatively small
+portion of the entire program: mainly the function \cd{list\_files}
+and its call site.
 
-Much of the above code is unaffected by staging; functions like
-\cd{sort\_by}, \cd{human\_size}, and even \cd{fileinfo} are oblivious
-to the stage in which they are run.  Therefore, the staged code comprises a
-relatively small portion of the entire program: mainly the function
-\cd{list\_files} and its call site.
-
-To understand the staging technique used here, it may help to examine
-the output of just the first stage of computation.  The following
-listing is cleaned up from the \MOC pretty-printer, with references to
-compiled values resolved in-place.
-
-\lgrindfile{dir-out.ml}
-
+To understand the staging technique, it may help to examine the output
+of just the first stage of computation.  The code block in
+figure~\ref{fig:dir-out.ml} is cleaned up from the \MOC
+pretty-printer, with references to compiled values substituted
+in-place.  
+%%
+\mylst{dir-out.ml}{Generated code for directory browser; see also
+  appendix~\ref{sec:dir.meta}.} 
+%%
 The \cd{list} is formed by testing filenames (in reverse alphabetical
@@
 
-\screenshot{gc}{Garbage collector statistics.  The descriptions are
-  copied directly from the \id{gc.mli} interface in OCaml.}
+\screenshot{gc}{Garbage collector statistics, from the OCaml \id{gc}
+  module.}
 
@@
 The implementation consists of two parts: a translator and a web
-server.  The translator, implemented in Standard ML,\footnote{Mainly
-  because the author has considerable experience with the ML-Yacc that
-  comes with SML/NJ, and is reluctant to learn the OCaml version!}
-transforms the server page syntax into plain \MOC, as illustrated in
-listing~\ref{lst:trans.ml}.  It recognizes all the server page blocks
-defined in section~\ref{sec:design} but does not completely parse the
-\MOC code contained within them.  Instead, it recognizes just enough
-of the keywords and delimiters to decide whether to add semi-colons or
-`{\KWfont in}' after each block.  This approach makes the translator
-fairly robust to minor changes in OCaml syntax.
+server.  The translator transforms the server page syntax into plain
+\MOC, as illustrated in figure~\ref{fig:trans.ml}.  It recognizes all
+the server page blocks defined in section~\ref{sec:design} but does
+not completely parse the \MOC code contained within them.  Instead, it
+recognizes just enough of the keywords and delimiters to decide
+whether to add semi-colons or `{\KWfont in}' after each block.  This
+approach makes the translator fairly robust to minor changes in OCaml
+syntax.
 
@@
 
-\lgrindfile{codeHandler.mli}
+\lgrindfile{codeHndl.mli}
 %%
@@
 
-The \cd{map} type maps a string (the URI) to a pair containing a
-function to (re-)generate the page (by running the first stage) and a
-ref cell where the page is cached.  No \cd{code} types show up here,
-but they are used within the \cd{unit\to page} function.  This
-function is implemented using the \Mrun{} (run) operator.  Passing
-around code values themselves is difficult because they must remain
-polymorphic to run them, and OCaml does not support rank-2
-polymorphism.
+Values of type \cd{map} tell the server which pages are mapped to
+which URIs.  Ideally, it would be defined as \cd{page code
+  StringMap.t}.  This is not currently possible: the \cd{page code}
+needs to be run, but in the current implementation of \MOC, the
+\Mrun{} operator cannot be used from within functions.  Passing around
+code values themselves has limited utility because they must remain
+polymorphic to run them, and OCaml does not yet support rank-2
+polymorphism.  (Although recent work by \citet{garrigue99semi} leads
+in that direction.)  Our work-around involves mapping the URI instead
+to a pair containing a function (defined at the top level) to generate
+the page (by running the first stage and then using the \Mrun{}
+operator on the generated code) and a ref cell where the resulting
+page is cached.
 
@@
 \label{sec:perf}
-One of the benefits of a \emph{compiled} implementation of staged
-programming---even byte-code compiled, like \MOC---is that it permits
-meaningful measurement.  Here, we describe the performance
+In this section, we describe the performance
 characteristics of the prototype, focusing in particular on the
 benefits of staging various web services.
-
+%%
 The single most important metric for evaluating web server performance
@@
 per unit of time.  To establish a baseline, we first tested the
-throughput of the Caml prototype HTTP server delivering chunks of
+throughput of our custom OCaml HTTP server delivering chunks of
 static data of various sizes, up to 64k bytes.
 %%
-Measurements were taken on an otherwise idle 1.8GHz Intel
-Xeon\texttrademark{} workstation\footnote{with 512kB cache, 768MB RAM,
-  and Ultra160 SCSI} running Linux 2.4.  We used \texttt{ab},
-the Apache HTTP server benchmarking tool, to issue requests
-from 8 threads simultaneously for 30 seconds.\footnote{\texttt{ab}
-  version 1.3d, available from \url{http://httpd.apache.org/}, 
-  invoked like this:
-  \texttt{ab -k -t 30 -c 8 }\textit{url}}
+Measurements were taken on an otherwise idle 1.8GHz Intel Xeon
+workstation\footnote{with 512kB cache, 768MB RAM, and Ultra160 SCSI}
+running Linux 2.6.  We used \texttt{ab}, the Apache HTTP server
+benchmarking tool, to issue requests from 8 threads simultaneously for
+30 seconds.\footnote{invoked like this: \texttt{ab -k -t 30 -c 8
+  }\textit{url}}
 
@@
 
-The baseline results are shown in figure~\ref{fig:static}.  The series
-marked “Caml file” was served by the \texttt{FileHandler} module.  For
-the “Caml code” series, the same files were treated as \MOC Server
-Pages, and thus turned into one big \texttt{puts} statement, to be
-executed by the \texttt{CodeHandler}.  The only reason the code
-wins in the beginning is that all code pages are already loaded into
-the server’s memory on startup (to work around a limitation of the
-prototype---see section~\ref{sec:impl}), but the \texttt{FileHandler}
-must read files from disk each time.
+The baseline results are shown in figure~\ref{fig:static}.  In the
+``OCaml FileHandler'' series, the files were treated just as static
+files, opened on disk, and copied out to the socket.  There are two
+sets of results for the FileHandler: one compiled as native code, and
+the other compiled as byte code.  A native code compiler for MetaOCaml
+is not yet available.  Therefore, most comparisons in this section
+will rely on byte code.  The native code results we are able to obtain
+at this time suggest how much improvement we can expect once native
+code is an option.
+
+For the ``OCaml CodeHandler'' series, the same set of files were
+treated as \MOC Server Pages, and thus translated (in advance) into
+one big \texttt{puts} statement, to be executed (as byte code) by the
+\texttt{CodeHandler} module.  The only reason the code beats the (byte
+code) FileHandler in the beginning is that all code pages are already
+loaded into the server's memory on startup (to work around a
+limitation of the prototype---see section~\ref{sec:impl}), but the
+FileHandler must read files from disk each time.
 
@@
 popular server-side computation system.\footnote{libphp4.so
-  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
+  (version 4.3.10) loaded into Apache 1.3} Here, we gave the same
 static data files the extension \texttt{.php}, so that Apache
@@
 
-Using the same methodology, we now consider the performance of
-the staged and unstaged power functions; see
-figure~\ref{fig:power}.  The throughput for the staged pages is
-about 30\% higher in the beginning, but as the exponents
-increase (again, note the log scale in the graph) the gap
-narrows.  In this program, staging removes the loop overhead,
-but the number of multiplications performed is the same either
-way.  Eventually, the cost of those multiplications dominates
-everything else.\footnote{The final result,
-  2\textsuperscript{8191} has 2,466 digits in base 10!}
+Using the same methodology, we now consider the performance of the
+staged and unstaged power functions; see figure~\ref{fig:power}.
+Since we are comparing staged vs. unstaged (and not native vs. byte
+code), these results need to be interpreted in two distinct groups.
+The top two lines are native code, while the bottom three lines are
+byte code.  Start with the byte code.
 
 \gnuplot{power}{Staged versus unstaged power function.  The
-  server page computes 2 to the given power
-  (2\textsuperscript{127}, 2\textsuperscript{255}, etc.)}
+  server computes $2^x$.}
+
+The throughput for the pages staged with MetaOCaml is about 30\%
+higher than the unstaged version in the beginning, but as the
+exponents increase (again, note the log scale in the graph) the gap
+narrows.  In this program, staging removes the loop overhead, but the
+cost of the multiplications and conversion to a decimal string (which
+involves non-trivial divisions) are needed either way.  Eventually,
+the cost of those operations dominates everything else.\footnote{The
+final result,
+  2\textsuperscript{8191} has 2,466 digits in base 10.}
+
+\mylst{powerhof.ml}{Staging the power function using higher-order
+functions.}
+
+At the suggestion of an anonymous reviewer, we also tried staging
+using higher-order functions in OCaml instead of the code splicing
+features of \MOC.  The relavant fragment of code is shown in
+figure~\ref{fig:powerhof.ml}.  Using continuation-passing style, it
+recursively builds up a closure that represents the computation
+required to compute any number to the power \cd n.
+
+As one might expect, staging using higher-order functions is slightly
+worse than producing specialized code as in \MOC, but better, of
+course, than no staging at all.  Moreover, this version compiles
+easily to native code with the standard OCaml tools, and this improves
+its performance significantly.  We do have reason to hope, however,
+that the native \MOC version, once working, will ultimately perform
+the best.
 
 Finally, we look at the performance of staged and unstaged directory
-browsing; see figure~\ref{fig:browse}.  Here, we created directories
-containing fixed numbers of files with random data.  The average file
-size was 32k.  The \emph{x}~axis shows the number of files in each
-directory.
+browsing; see figures~\ref{fig:browse} and~\ref{fig:hofbrowse}.  Here,
+we created directories containing fixed numbers of files with random
+data.  The average file size was 32k.  The \emph{x}~axis shows the
+number of files in each directory.
+
+\gnuplot{browse}{Staged versus unstaged directory browsing, in \MOC
+  and PHP.}
+
+\gnuplot{hofbrowse}{Directory browsing staged with higher-order
+functions vs. \MOC.}
 
@@
 which is then run by the server.  With PHP we must, as noted in
-section~\ref{sec:intro}, manage the quoting and persistence issues
-by hand.  The staged PHP code contains gems like this:
-\begin{verbatim}
-<?= "<?\n" ?>
-<?= "\$list = unserialize(\"".
-    addcslashes(serialize($list),'"').
-    "\");\n" ?>
-<?= "?>\n" ?>
-\end{verbatim}
-where the serialize occurs in stage one and the unserialize in stage two.
-
-\gnuplot{browse}{Staged versus unstaged directory browsing, in \MOC
-  and PHP.}
+section~\ref{sec:intro}, manage the quoting and persistence 
+by hand.  (This program is the source of the horrible
+\texttt{serialize}/\texttt{unserialize} code shown on
+page~\pageref{php-gem}.
 
@@
 files, the unstaged programs barely answered 18 requests per second.
-They would certainly succumb to the Slashdot effect!
+They would certainly succumb to the Slashdot effect!  (And compilation
+to native code made essentially no difference.)
 
@@
 
+We also tried staging the directory service using higher-order
+functions, the same technique as described for the power function.
+Again, this did not quite match the performance of the \MOC version
+(when comparing byte-code to byte-code).  However, in
+figure~\ref{fig:hofbrowse}, we can also see the substantial difference
+that compilation to native code can make.  We expect that the native
+\MOC compiler will enable even better throughput.
+
 \section{Related and future work}
@@
 Most of the \emph{server page} systems embed programs within web pages
-using similar techniques; exemplars include JSP~\cite{sun-jsp},
-ASP~\cite{microsoft-asp}, PHP~\cite{bakken04php},
-MSP~\cite{elsman03web} (based on SML), and XCaml~\cite{baretta04xcaml}
-(based on OCaml).  Ours appears to be the first such language with
+using similar techniques; examples include JSP, ASP, PHP,
+MSP~\cite{elsman03web} (based on SML), and AS/XCaml (based on
+OCaml).\footnote{Application System Xcaml, by Alessandro Baretta.
+  \url{http://www.asxcaml.org/}}
+%%
+Ours appears to be the first such language with
 explicit support for \emph{staging} the computation.  We inherited the
@@
 \citet{elsman04typing}, \citet{wallace99haxml}, \citet{hosoya03xduce},
-and \citet{ohl04xhtml} leverage ML-like type systems to validate
-(X)HTML generators.  Integrating their ideas into \MOC server pages
-may permit validation of the generated document as well.
+and Ohl\footnote{XHTML module.
+  \url{http://theorie.physik.uni-wuerzburg.de/~ohl/xhtml/}} 
+%%
+leverage ML-like type systems to validate (X)HTML generators.
+Integrating their ideas into \MOC server pages may permit validation
+of the generated document as well.
 
@@
 code, not the original, embedded code.  Camlp4 is a flexible
-pre-processor capable of modifying the concrete syntax of
-OCaml~\cite{rauglaudre03camlp4} while maintaining usable error
-messages.  Formulating the server page syntax with Camlp4 would likely
-be an improvement over the current, ad-hoc translator.
+pre-processor capable of modifying the concrete syntax of OCaml while
+maintaining usable error messages.\footnote{Camlp4, by Daniel de
+  Rauglaudre.  \url{http://pauillac.inria.fr/caml/camlp4/}}
+Formulating the server page syntax with Camlp4 would likely be an
+improvement over the current, ad-hoc translator.
 
@@
 Second, code values remain polymorphic until they are \emph{run} with
-the \Mrun{} operator.  Therefore, they cannot be passed to functions
-or stored in data structures (at least until OCaml supports rank-2
-polymorphism).
+the \Mrun{} operator.  Although they may be passed to functions and
+stored in data structures, code values \emph{cannot be run} from
+within functions.  That is, the \Mrun{} operator can only be used at
+the top level (until OCaml supports rank-2 polymorphism), which
+severely restricts the kinds of multi-stage programs one can write.
+
+Another interesting avenue is to incorporate work on
+\emph{offshoring}~\cite{eckhardt04offshoring}.  An alternative
+\emph{run} construct translates generated code blocks to lower-level
+languages for improved performance.  In the domain of web services, it
+is conceivable that a \MOC server page could then generate (via
+offshoring) a specialized server page in a more mainstream language,
+such as PHP or JSP.
 
@@
 express user interactions more naturally, given the stateless nature
-of HTTP.  We suspect that these techniques are orthogonal to the
+of HTTP.  We believe that these techniques are orthogonal to the
 staging features provided by \MOC.
@@
 or flexibility.  Our ideas would likely have greater impact if they
-could be implemented as a module for a real server such as Apache or
-AOLserver.
+were implemented as a module for a real server such as Apache or
+AOLserver.  Beyond the concerns outlined earlier in this section, this
+is likely to be `just' an engineering effort.
 
@@
 
-Many thanks to the anonymous reviewers of the \MOC workshop, for
-helpful recommendations on both the work and the presentation.
+Many thanks to the anonymous reviewers for the \MOC workshop, and for
+this article, for many helpful and enlightening recommendations on
+both the work and the presentation.  Thanks to Walid Taha for
+organizing the workshop, and to the rest of the \MOC team for their
+development efforts.
 
 \bibliographystyle{abbrvnat}
-\raggedright
 \bibliography{refs,../../cal}
+\appendix
+%\twocolumn
+\section{The directory browser}
+\label{sec:dir.meta}
+%\raggedright
+\setcodefonts{11pt}{10pt}
+\LGnuminterval=5000
+\LGindent=8pt
+\gdef\lgcommentstart{\end{lgrind}\vskip-3pt}
+\def\lgcommentend{\vskip.5\baselineskip\begin{lgrind}}
+
+\begin{lgrind}\input dir.meta
+\relax\end{lgrind}
 

### 2007-04-16 paper/paper.tex
svn-r29: final? revision for SCP sent to Walid
darcs-hash:20070416223432-e5a07-285e748aea7af3d7a84fc751d85cfc848de4e182.gz

@@
 \label{sec:intro}
-
 Modern dynamic web sites support many features for user collaboration
@@
 \begin{quote}
-  The problem is, if you have any sort of complicated code
-  on your page, that's what kills people. I generally think
-  that sites are not dying because they fill their pipe, but
-  because they have poorly written code.
-
-  As a general rule, anybody who has a pretty good
-  understanding of web design, they've done a good job of
-  learning what information to cache. What information needs
-  to be pregenerated. So when you're actually loading a
+  Anybody who has a pretty good
+  understanding of web design [has] done a good job of
+  learning what information to cache [and] what needs
+  to be pre-generated.  So when you're actually loading a
   page, even if it's a complicated page that looks dynamic
@@
   doing is putting together a bunch of puzzle pieces that
-  have been pregenerated, and making the simplest, quickest
+  have been pre-generated, and making the simplest, quickest
   decisions they possibly can.~\cite{turner03slashdot}
@@
 the computation performed by a web service.  Indeed, web publishing is
-an application area that is naturally staged:\footnote{Nevertheless,
-  it seems the literature on meta-programming has not yet acknowledged
+an application area that is naturally staged:\footnote{The literature
+on meta-programming has yet to acknowledge
   web services~\cite{w3c-wsdl} as a potential application area,
-  although \citet{sheard01accomp} does mention \emph{mobile code.}} %
+  although \citet{sheard01accomp} mentions \emph{mobile code.}
+  Analysis of other related work is in section~\ref{sec:related}.} %
 \begin{enumerate}
@@
 
-\item The content is rendered within the user's
-  browser---the \emph{display} stage.
+\item The content is rendered within the user's browser---the
+  \emph{display} stage.\footnote{Increasingly in modern web
+    applications, the rendered content is
+    \emph{interactive}---responsive to user input without a round-trip
+    to the server.}
 \end{enumerate}
-At \emph{each} stage there is an opportunity for computation
-to take place.  Consider the example of a conference
-calendar, such as the one illustrated in
-figure~\ref{fig:confcal}.  After specifying your areas of
-interest (perhaps using the ACM classifications), the server
-delivers a table of matching conferences, with dates,
-locations, deadlines, and links to conference web sites.
-Events remain in the table until a few weeks after they
-occur, but deadlines that have passed are marked in red.
-You may click on any column header to change the sort order.
-The next time you visit, the server remembers your
-preferences.  Perhaps it even sends you email to remind you
-of upcoming submission and registration deadlines.
+At \emph{each} stage there is an opportunity for computation to take
+place.\footnote{\citet{normark02programmatic} recognized these three
+stages as binding times, calling them 
+\emph{generated, calculated,} and \emph{dynamic} documents,
+respectively.}
+Consider the example of a conference calendar, such as the one
+illustrated in figure~\ref{fig:confcal}.  After specifying your areas
+of interest (perhaps using the ACM classifications), the server
+delivers a table of matching conferences, with dates, locations,
+deadlines, and links to conference web sites.  Events remain in the
+table until a few weeks after they occur, but deadlines that have
+passed are marked in red.  You may click on any column header to
+change the sort order.  The next time you visit, the server remembers
+your preferences.  Perhaps it even sends you email to remind you of
+upcoming submission and registration deadlines.
 
@@
 anything be computed off-line (at the \emph{publish} stage)?
-Yes!  Since this page is probably part of a much larger site
+Yes: since this page is probably part of a much larger site
 (whose structure does not change every day), the menus and
@@
 languages, including
-JSP,\footnote{\url{http://java.sun.com/products/jsp/}}
-ASP,\footnote{\url{http://www.asp.net/}} and
-PHP.\footnote{\url{http://www.php.net/}} The Common Gateway Interface
+JSP,
+ASP, and
+PHP.  The Common Gateway Interface
 (CGI)\footnote{\url{http://www.w3.org/CGI/}} addresses the
@@
 (\emph{display} stage), probably due to the difficulty of securing an
-installed base of interpreters.  JavaScript and Java applets are
+installed base of interpreters.  JavaScript, Java, and Flash applets are
 notable exceptions.
@@
 Strictly speaking, these languages are not exclusively confined to the
-stages that I have indicated: Javascript can be run client-side, PHP
+stages that I have indicated: Javascript can be run server-side, PHP
 can be run off-line, and so on.  Nevertheless, migrating code between
@@
 
+In this example and throughout this paper, a {\BGfont sans-serif}
+font is used for embedded \MOC code, with {\KWfont bold sans} reserved
+for keywords and code delimiters.  A \texttt{typewriter} font is used
+for \MOC character strings.  The regular serif font is used for plain
+text and HTML within the server page, and for comments within the \MOC
+code blocks.
+
 A very common use of code blocks is to print out (i.e., send to the
@@
 type \cd{string}.  Alternatively, messages may be formatted with
-\cd{printf} by placing the format string immediately after the code
+\cd{sprintf} by placing the format string immediately after the code
 delimiter.\footnote{In a departure from the standard OCaml syntax for
-  \cd{printf}, the arguments are comma-delimited.  This way, fewer
+  \cd{sprintf}, the arguments are comma-delimited.  This way, fewer
   parentheses are required when using the escape or lift operators on
@@
 \MOC augments OCaml with just three annotations, to indicate how
-programs are to be staged.  Brackets \cd{\Mbegincode like
-  these\Mendcode} construct future-stage computation.  The code within
-is not executed \emph{now,} but just returned, as a code value that can be
-run later.
+programs are to be staged.  Brackets \cd{\Mbegincode e\Mendcode}
+construct future-stage computation.  The code within is not executed
+in the current stage of computation, but just returned as a code value
+that can be run later.
 
 Within brackets, the splice or escape operator \cd{\Msplice e} may
-appear.  It interrupts the code construction to execute the expression
+appear.  It interrupts the code construction to evaluate the expression
 \cd{e} (in the current stage) and splice its result into the
-future-stage computation.  Thus, \cd{e} is required to return code of
+future-stage computation.  Thus, \cd{e} is required to evaluate to code of
 the proper type.
@@
 The \cd{a b c} represent publish-stage arguments (the names are
-specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
+specified with \cd{pragma args}), whereas \cd{req} and \cd{puts} are
 (fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
@@
   string\to string option} looks up the string value of the request
-parameter with a given name.  Since \cd{req} and \cd{puts} are
+parameter with a given name.  The \cd{option} type constructor permits
+the function to return \cd{None} if there is no matching parameter in
+the HTTP request.  On line 6, \cd{arg} is defined as a short-cut for
+retrieving a parameter by name.  Since \cd{req} and \cd{puts} are
 serve-stage arguments, it is incorrect to use them in the
 \emph{publish} (first) stage, and indeed the \MOC type system prevents
-this.
-
-A few finer points to note: the translator discards newlines that
-immediately follow code blocks (otherwise,
-figure~\ref{fig:trans.ml} would be dotted with %
-\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
-semi-colon or the {\KWfont in} keyword to code blocks, as required
-(see, for example, lines 7, 11, and 12 in figure~\ref{fig:trans.ml}.)
-%%
-Finally, because the translator partially parses the OCaml code
-blocks, it is not confused by code delimiters within strings and
-comments, nor by other uses of \la{} and \ra{} as operators.
+this.  These, along with the \cd{lift} function defined near the top,
+are essentially primitives from the point of view of the server page
+code.%%
+%%
+\footnote{Here are a few finer points about the translator: it discards
+newlines that immediately follow code blocks (otherwise,
+figure~\ref{fig:trans.ml} would be dotted with
+\cd{puts}~\texttt{"\char"5C{}n"} 
+statements).  It automatically appends a semi-colon or the {\KWfont
+in} keyword to code blocks, as required (see, for example, lines 7,
+11, and 12 in figure~\ref{fig:trans.ml}.)  Finally, because the
+translator partially parses the OCaml code blocks, it is not confused
+by code delimiters within strings and comments, nor by other uses of
+\la{} and \ra{} as operators.}
 
@@
 should be predictable.  Below is another example using \cd{stat}, this
-time to display the size of the server log file.
+time to display the size of some text file on the server.
 With the serve-stage argument \texttt{unit}, we can specify whether
@@
 
-To stage this, the \cd{stat} and \cd{float\_of\_int} calls could be
-lifted into the declaration block, as with the permissions example.
-Furthermore, we may also use \cd{lift} and the splice operator to
-perform the divisions in advance, even though they are underneath the
-{\KWfont match} (which cannot happen until the serve stage):
-\lgrindfile{size2.meta}%
+If this text file does not change frequently (and reporting 
+outdated information is no problem), the \cd{stat} and
+\cd{float\_of\_int} calls could be lifted into the declaration block,
+as with the permissions example.  Furthermore, we may also use
+\cd{lift} and the splice operator to perform the divisions in advance,
+even though they are underneath the {\KWfont match} (which cannot
+happen until the serve stage): \lgrindfile{size2.meta}%
 
@@
 
-We will see examples of most of these blocks in section~\ref{sec:eg}
-and appendix~\ref{sec:dir.meta}.
+In these cases, the expressions should not have code types: \cd{a}
+simply has type \cd{unit}, \cd{b} has type string, etc.  This is the
+only difference between the \cd{\Mtilde} and \cd{!} variants; both are
+evaluated in the first stage.  We will see examples of most of these
+blocks in section~\ref{sec:eg} and appendix~\ref{sec:dir.meta}.
 
@@
 used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
-may appear anywhere in the code.
-
-With publish-stage arguments, the page may be instantiated in
+may appear anywhere in the code.  Any identifiers following \cd{args}
+in the \cd{pragma} declaration will become publish-stage parameters.
+
+With publish-stage parameters, the page may be instantiated in
 countless ways.  Continuing the count-down example, we could map the
-URI \id{/longcount} to a code block generated with an actual argument
-of 99, while \id{/shortcount} refers to code with the actual
-argument 9.  Both are generated from the same server page.
+URI \id{/longcount} to a code block generated with an argument of 99,
+while \id{/shortcount} refers to code with the argument 9.  Both are
+generated from the same server page.
 
@@
 To explore the expressiveness of this design, we now look at a series
-of web services, organized to comprise a small web site.\footnote{A
-  demo is available at \url{http://comsci.liu.edu/metaserv/}} To give
+of web services, organized to comprise a small web site.  To give
 the services a similar look, we developed a site-wide style sheet, and
@@
 the scalability of this application.  To derive the un-staged version
-for comparison, we simply convert \cd{\Mquote{\Mtilde}} and
-\cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
+for comparison, we simply replace \cd{\Mquote{\Mtilde}} and
+\cd{\Mquote{!}} blocks with plain \cd{\Mquote{}} and remove all other
 brackets and escapes from the code in figure~\ref{fig:power.meta}.
@@
 containing the names and attributes of (and hyperlinks to) all
-the files in the directory.  In Apache, the
+the files in whatever directory is specified by the URL.  In Apache, the
 \texttt{mod\_autoindex} module provides this feature.
 %%
-ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
+ViewCVS is a more
 complex example of the same idea; it allows remote users to browse a
@@
 The \cd{FileHandler} takes a file system path as a parameter, and then
-interprets URIs as file names relative to that path.  If such a file
-exists, it sends it verbatim to the client.  If not, it passes
+interprets each URI as a file name relative to that path.  If such a
+file exists, it sends it verbatim to the client.  If not, it passes
 the responsibility on to the next handler in the chain.
@@
 time and the code values are already in memory when the server starts
-accepting requests.  This is because \MOC is currently not capable of
-reading and writing code values to a file.
-
-Here are the types of the page code and the map, along with signatures
-for the \cd{CodeHandler} (and two variants):
+accepting requests.  This is because \MOC does not currently permit
+programs to read and write code values to a file.
+
+Here are the types of the page code and the map, along with the
+signature for the \cd{CodeHandler}:
 
@@
 \cd{page} is the type of the code that is constructed by each server
-page; recall that the page takes two serve-time arguments: one
+page; recall that the page takes two serve-time parameters: one
 encapsulating the HTTP request, the other is the \cd{puts} function
@@
 
-Values of type \cd{map} tell the server which pages are mapped to
-which URIs.  Ideally, it would be defined as \cd{page code
-  StringMap.t}.  This is not currently possible: the \cd{page code}
-needs to be run, but in the current implementation of \MOC, the
-\Mrun{} operator cannot be used from within functions.  Passing around
-code values themselves has limited utility because they must remain
-polymorphic to run them, and OCaml does not yet support rank-2
-polymorphism.  (Although recent work by \citet{garrigue99semi} leads
-in that direction.)  Our work-around involves mapping the URI instead
-to a pair containing a function (defined at the top level) to generate
-the page (by running the first stage and then using the \Mrun{}
-operator on the generated code) and a ref cell where the resulting
+Data structures of type \cd{map} tell the server which pages are
+mapped to which URIs.  The range of the map is a pair: the first
+component (of type \cd{unit\to page}) is a function that will re-run
+stage one computation (using the \Mrun{} operator internally); the
+second component (of type \cd{page ref}) is a cell where the latest
 page is cached.
+
+The \cd{unit\to page} formulation is a work-around for what might
+otherwise be expressed as \cd{page code}, and run from within the
+\cd{CodeHandler}.  Unfortunately, code values in MetaOCaml must remain
+polymorphic to run them; the type is actually \cd{('a, page) code},
+for all \cd{'a}.  This is traditionally problematic in ML: we can
+define a polymorphic data structure, but not a data structure
+containing polymorphic values.  The work-around we used captures the
+polymorphic code value in a closure, which is then added to the map.
+Another possibility is to use the limited form of explicit
+polymorphism supported in OCaml, with the syntax \cd{\{f : 'a. ('a,
+page) code\}}.\footnote{Thanks to an anonymous reviewer for pointing
+this out.}  To use code values more directly, we would need rank-2
+polymorphism; work by \citet{garrigue99semi} is headed in that
+direction.
 
@@
 
-For the ``OCaml CodeHandler'' series, the same set of files were
+For the ``\MOC CodeHandler'' series, the same set of files were
 treated as \MOC Server Pages, and thus translated (in advance) into
@@
 using higher-order functions in OCaml instead of the code splicing
-features of \MOC.  The relavant fragment of code is shown in
-figure~\ref{fig:powerhof.ml}.  Using continuation-passing style, it
+features of \MOC.  The relevant fragment of code is shown in
+figure~\ref{fig:powerhof.ml}.  It
 recursively builds up a closure that represents the computation
@@
 \texttt{serialize}/\texttt{unserialize} code shown on
-page~\pageref{php-gem}.
+page~\pageref{php-gem}.)
 
@@
 files, the unstaged programs barely answered 18 requests per second.
-They would certainly succumb to the Slashdot effect!  (And compilation
-to native code made essentially no difference.)
+They would certainly succumb to the Slashdot effect---and compilation
+to native code made essentially no difference.
 
 This is exactly the kind of page where the real work needs to be done
-in advance.  But that doesn't mean it needs to be a completely static
+in advance.  But that does not mean it needs to be a completely static
 page, either.  By carefully staging the computation, we gathered the
@@
 new kinds of code blocks based on them.
+
+There is much related work on using various features of modern
+programming languages to implement sophisticated web services.
+\citet{normark02programmatic} proposed writing web pages in Scheme
+using the Lisp Abstracted Markup Language (LAML), which essentially
+represents HTML documents as S-expressions.  He distinguished three
+different binding times, when the Scheme program could be evaluated:
+off-line, page access time, or browse time (client-side).  These
+correspond precisely to the three stages we identified in section 1,
+but his programs did not \emph{transcend} different stages.  It is
+possible to implement staged programming in Scheme using
+\cd{quasiquote}, \cd{unquote}, and \cd{eval}~\cite{davies01modal};
+this would lead to a similar capability, save for the difference
+between static and dynamic checking of the generated code.
+
+\citet{queinnec00influence} and \citet{hughes00generalising} observed
+that multi-page web services could be implemented more naturally using
+continuations and \cd{call/cc} (essentially, by treating the server
+and user as coroutines).  This way, we can treat the entire service as
+one program---that suspends itself while waiting for user
+input---instead of developing each page as a separate program.  This
+is a valuable technique, and appears to be orthogonal to staging;
+although it would be worth implementing both together to determine if
+there are unforeseen interactions.
+
+\citet{graunke01programming,graunke03modeling} took up this idea and
+mixed it with other language features: first-class modules, preemptive
+threads, and \emph{custodians,} for managing resource consumption.
+The result is a server that achieves lower overhead for dynamic
+services compared to Apache CGI (although the overhead of CGI is
+avoided by most modern dynamic services, by embedding interpreters for
+JSP, PHP, or similar in the server).  \citet{matthews04automatically}
+compile such direct-style interactive programs into CGI-style scripts.
 
@@
 Whenever we want to re-run the publish stage, we could do so
-independently of the server.  Two limitations of the \MOC
-implementation itself appear to stand in the way of this ideal.
-First, code values cannot at present be written to or read from files.
-Second, code values remain polymorphic until they are \emph{run} with
-the \Mrun{} operator.  Although they may be passed to functions and
-stored in data structures, code values \emph{cannot be run} from
-within functions.  That is, the \Mrun{} operator can only be used at
-the top level (until OCaml supports rank-2 polymorphism), which
-severely restricts the kinds of multi-stage programs one can write.
+independently of the server.
 
@@
 offshoring) a specialized server page in a more mainstream language,
-such as PHP or JSP.
+such as PHP, JSP, or---for client-side computation---JavaScript.
 
@@
 
-Many thanks to the anonymous reviewers for the \MOC workshop, and for
-this article, for many helpful and enlightening recommendations on
-both the work and the presentation.  Thanks to Walid Taha for
-organizing the workshop, and to the rest of the \MOC team for their
-development efforts.
+Sincere thanks to the anonymous reviewers, for many helpful and
+enlightening recommendations on both the work and the presentation.
+Thanks to Walid Taha for organizing the \MOC workshop, and to the rest
+of the \MOC team for their development efforts.
 
 \bibliographystyle{abbrvnat}
-\bibliography{refs,../../cal}
+\bibliography{refs,../../bib/library}
 \appendix

### 2007-04-16 paper/paper.tex
svn-r30: added citation to Takebe
darcs-hash:20070416223508-e5a07-36329b3c836c46dff2cb183c1198a6bfb29f939c.gz

@@
 where the \texttt{serialize} library function occurs in stage one and
-the \texttt{unserialize} in stage two.  Notice that the \verb+$+
+the \texttt{unserialize} in stage two.  Notice that the \verb+$+ %$
 preceding the first occurrence of \verb+list+ is quoted, but the
@@
 between static and dynamic checking of the generated code.
+
+\citet{takebe02caching} developed a partial evaluation technique for
+PHP.  Their tool, PHP-Mix, reads standard PHP source code and performs
+binding time analysis.  Then, it generates a semi-static script as
+output.  It understands a significant subset of the PHP function
+library, including the database connectivity.  Unfortunately, some
+properties of PHP---such as the lack of variable declarations and the
+single global scope---cause the analysis to be overly conservative,
+so it does not always achieve the optimizations we desire.
 
@@
 enlightening recommendations on both the work and the presentation.
+%%
+Thanks to my M.S. student, Kiichi Takeuchi, for exploring some
+applications of these ideas, and for helping me understand the work of
+\citet{takebe02caching} (written in Japanese).
+%%
 Thanks to Walid Taha for organizing the \MOC workshop, and to the rest
